### 2024-09-24 paper/paper.qmd
removed unused sections from paper.qmd
still needs to be updated with the relevant data

---
title: "Capital Budget Allocations and Median Household Income "
subtitle: "My subtitle if needed"
author: 
  - Maria Mangru 
thanks: "Code and data are available at: https://github.com/MariaMangru/Capital-Budget-Allocations-and-Median-Household-Income."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References

### 2024-09-24 scripts/06 - combine_data.R
created a new script to combine datasets

@@ -1,0 +1,14 @@
+#### Preamble ####
+# Purpose: Explores data in an attempt to understand it more
+# Author: Maria Mangru
+# Date: September 2024
+# Contact: maria.mangru@mail.utoronto.ca
+# License: MIT
+
+#### Workspace setup ####
+library(dplyr)
+library(tidyverse)
+library(readr)
+library(knitr)
+library(kableExtra)
+library(scales)

### 2024-09-24 scripts/06 - combine_data.R
begin combining datasets

@@ -12,3 +12,143 @@ library(readr)
 library(knitr)
 library(kableExtra)
 library(scales)
+
+
+### Load Data ###
+budget_data_clean <- read_csv("data/analysis_data/2022-2031-budget_data.csv")
+ward_data_clean <- read_csv("data/analysis_data/2023-WardProfile-Data.csv")
+
+
+### Combine Data ##
+# Rename the first column to 'Category' for clarity
+ward_profile_clean <- ward_data_clean |>
+  rename(Category = `...1`) |> 
+  select(-Toronto) |>            
+  pivot_longer(
+    cols = starts_with("Ward"),
+    names_to = "Ward Number",
+    values_to = "Value"
+  ) |>
+  pivot_wider(
+    names_from = Category,
+    values_from = Value
+  )
+
+View(ward_profile_clean)
+
+# Identify year columns
+year_cols <- as.character(2022:2031)
+
+# Aggregate total funding per ward
+total_funding_ward <- budget_data_clean %>%
+  group_by(`Ward Number`) %>%
+  summarise(
+    Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)
+  ) %>%
+  arrange(desc(Total_Funding))
+
+# Merge total_funding_ward with ward_data_clean
+combined_data <- total_funding_ward %>%
+  left_join(ward_data_clean, by = "Ward Number")
+
+
+# Calculate funding per capita
+combined_data <- combined_data %>%
+  mutate(
+    Funding_Per_Capita = Total_Funding / `Total Population`
+  )
+
+# Calculate funding as a percentage of median income
+combined_data <- combined_data %>%
+  mutate(
+    Funding_Per_Median_Income = Total_Funding / `Median total income of households in 2020 ($)`
+  )
+
+
+### Tables ### 
+# Combined Funding and Demographics Table
+combined_table <- combined_data %>%
+  select(
+    `Ward Number`,
+    Total_Funding,
+    `Total Population`,
+    `Average total income of households in 2020 ($)`,
+    Funding_Per_Capita,
+    Funding_Per_Median_Income
+  ) %>%
+  mutate(
+    Total_Funding = dollar(Total_Funding),
+    `Average Total Income` = dollar(`Average total income of households in 2020 ($)`),
+    Funding_Per_Capita = dollar(Funding_Per_Capita),
+    Funding_Per_Median_Income = dollar(Funding_Per_Median_Income)
+  )
+
+combined_table %>%
+  kable(
[…]

### 2024-09-24 scripts/06 - combine_data.R
kept combining data
needs lots of fixing :(

@@ -15,11 +15,11 @@ library(scales)
 
 
 ### Load Data ###
-budget_data_clean <- read_csv("data/analysis_data/2022-2031-budget_data.csv")
+X2022_2031_budget_data <- read_csv("data/analysis_data/2022-2031-budget_data.csv", col_types = cols(`Ward Number` = col_character()))
 ward_data_clean <- read_csv("data/analysis_data/2023-WardProfile-Data.csv")
 
 
-### Combine Data ##
+### Restructure ##
 # Rename the first column to 'Category' for clarity
 ward_profile_clean <- ward_data_clean |>
   rename(Category = `...1`) |> 
@@ -36,119 +36,48 @@ ward_profile_clean <- ward_data_clean |>
 
 View(ward_profile_clean)
 
-# Identify year columns
-year_cols <- as.character(2022:2031)
+# Function to clean dollar-formatted strings
+clean_dollar <- function(x){
+  as.numeric(str_replace_all(x, "[$,]", ""))
+}
 
-# Aggregate total funding per ward
-total_funding_ward <- budget_data_clean %>%
-  group_by(`Ward Number`) %>%
+# Function to clean percentage-formatted strings
+clean_percent <- function(x){
+  as.numeric(str_replace(x, "%", ""))
+}
+
+
+# Apply cleaning functions to appropriate columns
+ward_profile_clean <- ward_profile_clean |>
+  mutate_at(vars(contains("$")), clean_dollar) |>   
+  mutate_at(vars(contains("%")), clean_percent) |>
+  mutate_at(vars(-`Ward Number`), ~ as.numeric(.))    
+
+
+### Combine Data ###
+
+# Perform a left join to combine funding data with ward demographics
+combined_data <- X2022_2031_budget_data |>
+  left_join(ward_profile_clean, by = "Ward Number")
+
+combined_data <- combined_data |>
+  group_by(`Ward Number`) |>
   summarise(
     Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)
-  ) %>%
+  ) |>
   arrange(desc(Total_Funding))
 
-# Merge total_funding_ward with ward_data_clean
-combined_data <- total_funding_ward %>%
-  left_join(ward_data_clean, by = "Ward Number")
+colnames(ward_profile_clean)
 
-
-# Calculate funding per capita
-combined_data <- combined_data %>%
+# Calculate Funding Per Capita
+combined_data <- combined_data |>
   mutate(
-    Funding_Per_Capita = Total_Funding / `Total Population`
[…]

### 2024-09-24 scripts/06 - combine_data.R
updated code for combined data

@@ -15,69 +15,57 @@ library(scales)
 
 
 ### Load Data ###
-X2022_2031_budget_data <- read_csv("data/analysis_data/2022-2031-budget_data.csv", col_types = cols(`Ward Number` = col_character()))
-ward_data_clean <- read_csv("data/analysis_data/2023-WardProfile-Data.csv")
+# Budget funding data
+budget_data_clean <- read_csv("data/analysis_data/2022-2031-budget_data.csv")
+
+# Demographic data
+demographic_data <- read_csv("data/analysis_data/demographic_data.csv")
+
+# Ensure that `Ward Number` is consistent in both datasets
+budget_data_clean <- budget_data_clean |>
+  mutate(`Ward Number` = as.integer(`Ward Number`))
+
+demographic_data <- demographic_data |>
+  rename(`Ward Number` = `...1`) |>
+  mutate(`Ward Number` = as.integer(`Ward Number`))
 
 
-### Restructure ##
-# Rename the first column to 'Category' for clarity
-ward_profile_clean <- ward_data_clean |>
-  rename(Category = `...1`) |> 
-  select(-Toronto) |>            
+### Restructure ###
+# Reshape budget data to long format 
+year_cols <- as.character(2022:2031)
+
+budget_data_long <- budget_data_clean %>%
   pivot_longer(
-    cols = starts_with("Ward"),
-    names_to = "Ward Number",
-    values_to = "Value"
-  ) |>
-  pivot_wider(
-    names_from = Category,
-    values_from = Value
+    cols = all_of(year_cols),
+    names_to = "Year",
+    values_to = "Funding"
   )
 
-View(ward_profile_clean)
-
-# Function to clean dollar-formatted strings
-clean_dollar <- function(x){
-  as.numeric(str_replace_all(x, "[$,]", ""))
-}
-
-# Function to clean percentage-formatted strings
-clean_percent <- function(x){
-  as.numeric(str_replace(x, "%", ""))
-}
-
-
-# Apply cleaning functions to appropriate columns
-ward_profile_clean <- ward_profile_clean |>
-  mutate_at(vars(contains("$")), clean_dollar) |>   
-  mutate_at(vars(contains("%")), clean_percent) |>
-  mutate_at(vars(-`Ward Number`), ~ as.numeric(.))    
-
-
-### Combine Data ###
-
-# Perform a left join to combine funding data with ward demographics
-combined_data <- X2022_2031_budget_data |>
[…]

### 2024-09-24 scripts/05 - budget_data_results.R
removed combine data file

@@ -22,7 +22,7 @@ year_cols <- as.character(2022:2031)
 
 
 # Reshape data to long format
-budget_data_long <- budget_data_clean %>%
+budget_data_long <- budget_data_clean |>
   pivot_longer(
     cols = all_of(year_cols),
     names_to = "Year",

### 2024-09-24 scripts/06 - combine_data.R

@@ -1,71 +1,0 @@
-#### Preamble ####
-# Purpose: Explores data in an attempt to understand it more
-# Author: Maria Mangru
-# Date: September 2024
-# Contact: maria.mangru@mail.utoronto.ca
-# License: MIT
-
-#### Workspace setup ####
-library(dplyr)
-library(tidyverse)
-library(readr)
-library(knitr)
-library(kableExtra)
-library(scales)
-
-
-### Load Data ###
-# Budget funding data
-budget_data_clean <- read_csv("data/analysis_data/2022-2031-budget_data.csv")
-
-# Demographic data
-demographic_data <- read_csv("data/analysis_data/demographic_data.csv")
-
-# Ensure that `Ward Number` is consistent in both datasets
-budget_data_clean <- budget_data_clean |>
-  mutate(`Ward Number` = as.integer(`Ward Number`))
-
-demographic_data <- demographic_data |>
-  rename(`Ward Number` = `...1`) |>
-  mutate(`Ward Number` = as.integer(`Ward Number`))
-
-
-### Restructure ###
-# Reshape budget data to long format 
-year_cols <- as.character(2022:2031)
-
-budget_data_long <- budget_data_clean %>%
-  pivot_longer(
-    cols = all_of(year_cols),
-    names_to = "Year",
-    values_to = "Funding"
-  )
-
-# Summarize total funding per ward
-total_funding_ward <- budget_data_clean %>%
-  group_by(`Ward Number`) %>%
-  summarise(
-    Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)
-  ) %>%
-  arrange(desc(Total_Funding))
-
-# Summarize total funding per ward and per program/agency
-total_funding_program_ward <- budget_data_clean %>%
-  group_by(`Ward Number`, `Program/Agency Name`) %>%
-  summarise(Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)) %>%
-  arrange(`Ward Number`, desc(Total_Funding))
-
-# Merge total funding with demographic data
-funding_demographics <- total_funding_ward %>%
-  left_join(demographic_data, by = "Ward Number")
-
-# Inspect the merged data
-head(funding_demographics)
-
-
-
-
-
-
-
-

### 2024-09-25 scripts/merge_test.R
attempt to merge budget data and demographic data

@@ -1,0 +1,106 @@
+#### Workspace Setup ####
+library(dplyr)
+library(tidyr)
+library(readr)
+library(knitr)
+library(kableExtra)
+library(scales)
+library(stringr)
+library(janitor)
+library(ggplot2)
+
+### Data Handling ###
+
+# Read in the budget data
+budget_data <- read_csv("data/analysis_data/budget.csv")
+
+# Read in the demographic data
+demographic_data <- read_csv("data/analysis_data/demographic.csv")
+
+# View the structure of the datasets
+str(budget_data)
+str(demographic_data)
+
+
+# Identify year columns in budget data
+year_cols <- as.character(2022:2031)
+
+# Ensure that year columns and ward number are numeric
+budget_data <- budget_data |>
+  mutate(across(all_of(c(year_cols, "Total 10 Year", "Ward Number")), as.numeric))
+
+# Aggregate total funding per ward number
+total_funding_ward <- budget_data |>
+  group_by(`Ward Number`) |>
+  summarise(
+    total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
+  ) |>
+  arrange(desc(total_funding))
+
+# View total funding per ward
+print(total_funding_ward)
+
+# Process demographic data
+# Clean column names
+demographic_data <- demographic_data %>%
+  clean_names()
+
+# Pivot demographic data from wide to long format
+demographic_long <- demographic_data %>%
+  pivot_longer(
+    cols = -Category,
+    names_to = "Ward Number",
+    values_to = "Value"
+  )
+
+
+# Pivot demographic data back to wide format to get one row per ward
+demographic_wide <- demographic_long %>%
+  pivot_wider(
+    names_from = Category,
+    values_from = Value
+  )
+
+demographic_wide$`Ward Number` <- as.numeric(as.character(demographic_wide$`Ward Number`))
+
+
+as.numeric(as.character(total_funding_ward$`Ward Number`))
+
+sapply(total_funding_ward, class)
+sapply(demographic_wide, class)
+
+# Merge budget data with demographic data
+merged_data <- total_funding_ward %>%
+  left_join(demographic_wide, by = "Ward Number")
+
+
+colnames(merged_data)
+
+
+
+### Tables and Visualizations ###
+
+# Select relevant columns for the table
+merged_table <- merged_data %>%
+  select(
+    `Ward Number`,
+    `total_funding`,
+    `Total Population`,
+    `Median total income of households in 2020 ($)`,
[…]

### 2024-09-25 paper/paper.qmd
began draft of abstract and introduction

@@
 ---
-title: "Capital Budget Allocations and Median Household Income "
-subtitle: "My subtitle if needed"
+title: "Capital Budget Funding in Toronto"
+subtitle: "An Analysis of the Funding Disparties Across Toronto Wards (2022 - 2031)"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This research examines the allocation of capital funding across Toronto's 25 wards and explores how these allocations correspond with various socioeconomic indicators.
+The analysis reveals significant disparities in per capita funding, with higher-income wards often receiving more funding than lower-income wards."
 format: pdf
@@
 
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-library(tidyverse)
-library(palmerpenguins)
-```
-
 # Introduction
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Toronto, often referred to as a cultural mosaic, is home to a diverse population with roots from all around the world.
+It is also city where its neighbourhoods vary significantly in income, educational attainment, and demographic makeup.
+The allocation of capital funding plays an important role in shaping the infrastructure and services available to residents.
+This impacts their quality of life and access to opportunities.
 
-The remainder of this paper is structured as follows. @sec-data....
+Despite the importance of equitable funding, there is evidence that budget distributions may not align with the needs of different communities.
+\[Write about these two:https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8186789/ , http://neighbourhoodchange.ca/documents/2016/05/walks-etal-2016-income-inequality-toronto-york-rp-238.pdf\]
+
+By analyzing data from the City of Toronto's Capital Budget & Plan, as well as its Ward Profiles, this research finds disparities in per capita funding among the 25 wards.
+The findings indicate that higher-income wards tend to receive more funding per capita than lower-income wards.
+
+The remainder of the paper is organized as follows: Section \ref{sec-data} describes the data sources and methodology used in the analysis.
+Section \ref{sec-results} presents the results, highlighting key findings.
+Section \ref{sec-discussion} discussed the implications of the findings as well as limitations and recommendations for next steps.
 
@@
 Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
 
@@
 
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
-
-# Results
+# Results {#sec-results}
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
-
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-09-25 paper/paper.qmd
wrote about the data sources used

@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+## Data Sources
 
-Talk more about it.
+This research utilized two primary datasets provided by the City of Toronto's Open Data Portal.
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+1.  Capital Budget & Plan By Ward (2022-2031): \\
+This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
+It includes funding allocations for various programs such as Children's Services, Parks, Foresty & Recreation, Transportation Services, and others.
+The data provides a detailed breakdown of how much funding each ward is set to receive for specific projects and services.
+
+2.  Ward Profiles (2021): \\
+This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
+It was filtered to remain the variables which were most useful for this research.
+These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population.
+[may need to include others, depends on what I decide to go with in the end.]
 
@@
 
-Our results are summarized in @tbl-modelresults.
+Lots of tables and graphs to be included here once they're properly formatted!
 
@@
 ## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-Weaknesses and next steps should also be included.
-
 \newpage

### 2024-09-25 paper/paper.qmd
included table showing yearly and total funding for ward

@@
 ---
+
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+
+#### Workspace setup ####
+library(dplyr)
+library(tidyverse)
+library(readr)
+library(knitr)
+library(kableExtra)
+library(scales)
+
+### Data handling ###
+# Import cleaned data 
+budget_data_clean <- read_csv("../data/analysis_data/2022-2031-budget_data.csv")
+
+# Identify year columns
+year_cols <- as.character(2022:2031)
+
+# Reshape data to long format
+budget_data_long <- budget_data_clean |>
+  pivot_longer(
+    cols = all_of(year_cols),
+    names_to = "Year",
+    values_to = "Funding"
+  )
+
+### Summing funding values ###
+
+# Total overall funding per ward number 
+total_funding_ward <- budget_data_clean |>
+  group_by(`Ward Number`) |>
+  summarise(
+    Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)
+  ) |>
+  arrange(desc(Total_Funding))
+
+# Funding for each year per ward number 
+funding_yearly_ward <- budget_data_long |>
+  group_by(`Ward Number`, Year) |>
+  summarise(Funding = sum(Funding, na.rm = TRUE)) |>
+  arrange(`Ward Number`, Year)
+
+# Total funding per program/agency per ward number 
+total_funding_program_ward <- budget_data_clean |>
+  group_by(`Ward Number`, `Program/Agency Name`) |>
+  summarise(Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)) |>
+  arrange(`Ward Number`, desc(Total_Funding))
+
+# Yearly funding per program/agency per ward number - maybe remove later
+funding_yearly_program_ward <- budget_data_long |>
+  group_by(`Ward Number`, `Program/Agency Name`, Year) |>
+  summarise(
+    Funding = sum(Funding, na.rm = TRUE)
+  ) |>
+  arrange(`Ward Number`, `Program/Agency Name`, Year)
+
+```
 
@@
 
-1.  Capital Budget & Plan By Ward (2022-2031): \\
-This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
+1.
+Capital Budget & Plan By Ward (2022-2031): \\ This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
 It includes funding allocations for various programs such as Children's Services, Parks, Foresty & Recreation, Transportation Services, and others.
@@
 
-2.  Ward Profiles (2021): \\
-This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
+2.
+Ward Profiles (2021): \\ This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
 It was filtered to remain the variables which were most useful for this research.
 These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population.
-[may need to include others, depends on what I decide to go with in the end.]
+\[may need to include others, depends on what I decide to go with in the end.\]
 
@@
 Lots of tables and graphs to be included here once they're properly formatted!
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+# Yearly Funding per Ward Number with Total Column
+funding_yearly_ward_table <- funding_yearly_ward |>
+  pivot_wider(names_from = Year, values_from = Funding, values_fill = 0) |> 
+  rowwise() |> 
+  mutate(`Total Funding` = sum(c_across(all_of(year_cols)))) |> 
+  ungroup() |> 
+  mutate_at(vars(all_of(year_cols), `Total Funding`), ~ dollar(.)) |> 
+  arrange(`Ward Number`)
+
+funding_yearly_ward_table |> 
+  kable(
+    col.names = c("Ward Number", year_cols, "Total Funding"),
+    caption = "Yearly Funding for Wards"
+  ) |>
+
+  kableExtra::kable_styling(bootstrap_options = "condensed", full_width = F, font_size = 6, position = "center")
+
+```
 

### 2024-09-25 scripts/05 - budget_data_results.R
small formatting changes to yearly funding graph in quarto

@@ -109,7 +109,6 @@ total_funding_program_ward_table <- total_funding_program_ward |>
 total_funding_program_ward_table |>
   kable(
     col.names = c("Ward Number", "Program/Agency Name", "Total Funding"),
-    format = "html",
     caption = "Total Funding per Program/Agency Name for Each Ward Number"
   ) |>
   kable_styling(full_width = FALSE, position = "center") |>
@@ -124,7 +123,6 @@ funding_yearly_program_ward_table <- funding_yearly_program_ward |>
 funding_yearly_program_ward_table |>
   kable(
     col.names = c("Ward Number", "Program/Agency Name", year_cols),
-    format = "html",
     caption = "Yearly Funding per Program/Agency Name"
   ) |>
   kable_styling(full_width = TRUE, position = "center") |>
@@ -174,7 +172,7 @@ ggplot(total_funding_program_ward,
        aes(x = reorder(`Program/Agency Name`, Total_Funding), y = Total_Funding, fill = `Program/Agency Name`)) +
   geom_bar(stat = "identity") +
   labs(
-    title = "Total Funding per Top 5 Program/Agency Names for Each Ward",
+    title = "Total Funding by Program/Agency Names for Each Ward",
     x = "Program/Agency Name",
     y = "Total Funding"
   ) +

### 2024-09-25 scripts/05 - budget_data_results.R
removed lines of code not needed

@@ -154,7 +154,7 @@ ggplot(funding_yearly_ward, aes(x = as.integer(Year), y = Funding, color = `Ward
     title = "Funding Trends Over Years",
     x = "Year",
     y = "Funding",
-    color = "Ward Number"
+    #color = "Ward Number"
   ) +
   scale_x_continuous(breaks = 2022:2031) +
   scale_y_continuous(labels = dollar_format()) +
@@ -165,25 +165,6 @@ ggplot(funding_yearly_ward, aes(x = as.integer(Year), y = Funding, color = `Ward
     axis.title = element_text(size = 14),
     legend.title = element_text(size = 13),
     legend.text = element_text(size = 11)
-  )
-
-# Total Funding by Program/Agency Names for Each Ward Number - NEEDS LOTS OF FIXING!! CANNOT READ
-ggplot(total_funding_program_ward, 
-       aes(x = reorder(`Program/Agency Name`, Total_Funding), y = Total_Funding, fill = `Program/Agency Name`)) +
-  geom_bar(stat = "identity") +
-  labs(
-    title = "Total Funding by Program/Agency Names for Each Ward",
-    x = "Program/Agency Name",
-    y = "Total Funding"
-  ) +
-  scale_y_continuous(labels = dollar_format()) +
-  theme_minimal() +
-  facet_wrap(~ `Ward Number`, scales = "free_y") +
-  theme(
-    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
-    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
-    axis.title = element_text(size = 14),
-    legend.position = "none"
   )
 
 

### 2024-09-25 scripts/merge_test.R

@@ -1,3 +1,13 @@
+#### Preamble ####
+# Purpose: Merge demographic and budget data 
+# Author: Maria Mangru
+# Date: September 2024
+# Contact: maria.mangru@mail.utoronto.ca
+# License: MIT
+
+
+
+
 #### Workspace Setup ####
 library(dplyr)
 library(tidyr)
@@ -12,10 +22,10 @@ library(ggplot2)
 ### Data Handling ###
 
 # Read in the budget data
-budget_data <- read_csv("data/analysis_data/budget.csv")
+budget_data <- read_csv("data/analysis_data/20221-2031-budget_data.csv")
 
 # Read in the demographic data
-demographic_data <- read_csv("data/analysis_data/demographic.csv")
+demographic_data <- read_csv("data/analysis_data/2023-WardProfile-Data.csv")
 
 # View the structure of the datasets
 str(budget_data)
@@ -25,9 +35,6 @@ str(demographic_data)
 # Identify year columns in budget data
 year_cols <- as.character(2022:2031)
 
-# Ensure that year columns and ward number are numeric
-budget_data <- budget_data |>
-  mutate(across(all_of(c(year_cols, "Total 10 Year", "Ward Number")), as.numeric))
 
 # Aggregate total funding per ward number
 total_funding_ward <- budget_data |>
@@ -40,13 +47,9 @@ total_funding_ward <- budget_data |>
 # View total funding per ward
 print(total_funding_ward)
 
-# Process demographic data
-# Clean column names
-demographic_data <- demographic_data %>%
-  clean_names()
 
 # Pivot demographic data from wide to long format
-demographic_long <- demographic_data %>%
+demographic_long <- demographic_data |>
   pivot_longer(
     cols = -Category,
     names_to = "Ward Number",
@@ -55,7 +58,7 @@ demographic_long <- demographic_data %>%
 
 
 # Pivot demographic data back to wide format to get one row per ward
-demographic_wide <- demographic_long %>%
+demographic_wide <- demographic_long |>
   pivot_wider(
     names_from = Category,
     values_from = Value
@@ -70,7 +73,7 @@ sapply(total_funding_ward, class)
 sapply(demographic_wide, class)
 
 # Merge budget data with demographic data
-merged_data <- total_funding_ward %>%
+merged_data <- total_funding_ward |>
   left_join(demographic_wide, by = "Ward Number")
 
 
@@ -81,25 +84,25 @@ colnames(merged_data)
 ### Tables and Visualizations ###
 
 # Select relevant columns for the table
[…]

### 2024-09-26 scripts/03 - data_cleaning.R
updated data cleaning code to remove NAs

@@ -1,7 +1,7 @@
 #### Preamble ####
-# Purpose: Cleans the raw data downloaded into a format which can be analyzed
+# Purpose: Cleans the raw data downloaded, into a format which can be analyzed
 # Author: Maria Mangru
-# Date: September 2024
+# Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
 
@@ -9,28 +9,32 @@
 library(dplyr)
 library(tidyverse)
 library(readr)
+library(stringr)
 
 
 #### Data Cleaning ####
 
 ## Ward Profile Data ## 
+
 # Clean ward profile 
 ward_profile <- read_csv("data/raw_data/2023-WardProfiles-2011-2021-CensusData.csv")
 
-# Remove trailing whitespaces from character columns
-ward_profile_clean <- ward_profile |>
-  mutate(across(where(is.character), str_trim))
+# Remove rows with NAs
+ward_profile <- na.omit(ward_profile)
 
 # Save cleaned ward profile data 
-write_csv(ward_profile_clean, "data/analysis_data/2023-WardProfile-Data.csv")
+write_csv(ward_profile, "data/analysis_data/clean_ward_demo_data.csv")
+
 
 
 ## 2022 - 2031 Budget Data ##
+
 # Clean budget data 
 budget_data <- read_csv("data/raw_data/2022-2031-capital-budget-and-plan-details.csv")
 
 # Remove rows where 'Ward Number' is "CW"
-filtered_budget_data <- budget_data[-(which(budget_data$'Ward Number' %in% "CW")),]
+budget_data <- budget_data |>
+  filter(`Ward Number` != "CW")
 
 # Identify year columns
 year_cols <- as.character(2022:2031)
@@ -48,6 +52,6 @@ budget_data_clean <- filtered_budget_data |>
 
 
 # Save cleaned budget data 
-write_csv(budget_data_clean, "data/analysis_data/2022-2031-budget_data.csv")
+write_csv(budget_data_clean, "data/analysis_data/clean_budget_data.csv")
 
 

### 2024-09-26 scripts/03 - data_cleaning.R
updated data cleaning

@@ -47,7 +47,7 @@ clean_funding <- function(x) {
 }
 
 # Apply the cleaning function to year columns
-budget_data_clean <- filtered_budget_data |>
+budget_data_clean <- budget_data |>
   mutate(across(all_of(year_cols), clean_funding))
 
 

### 2024-09-26 scripts/05-analysis.R

@@ -1,0 +1,6 @@
+#### Preamble ####
+# Purpose: Performs analysis of budget and merged data
+# Author: Maria Mangru
+# Date: 27 September 2024
+# Contact: maria.mangru@mail.utoronto.ca
+# License: MIT

### 2024-09-26 scripts/05-analysis.R
combined scripts for budget data and merge test into one script, updated the graphs and tables

@@ -4,3 +4,262 @@
 # Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
+
+#### Workspace setup ####
+library(dplyr)
+library(tidyr)
+library(readr)
+library(knitr)
+library(kableExtra)
+library(scales)
+library(ggplot2)
+library(stringr)
+library(janitor)
+
+
+### Data Handling ###
+
+# Read in the cleaned budget data
+budget_data <- read_csv("data/analysis_data/clean_budget_data.csv")
+
+# Read in the cleaned demographic data
+demographic_data <- read_csv("data/analysis_data/clean_ward_demo_data.csv")
+
+# Identify year columns in budget data
+year_cols <- as.character(2022:2031)
+
+# Aggregate total funding per ward number
+total_funding_ward <- budget_data |>
+  group_by(`Ward Number`)|>
+  summarise(
+    total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
+  ) |>
+  arrange(desc(total_funding))
+
+# Adjust total funding (amounts are in thousands of dollars)
+total_funding_ward <- total_funding_ward |>
+  mutate(total_funding = total_funding * 1000)
+
+# Pivot demographic data from wide to long format
+demographic_long <- demographic_data |>
+  pivot_longer(
+    cols = -category,
+    names_to = "Ward Number",
+    values_to = "Value"
+  )
+
+# Pivot demographic data back to wide format to get one row per ward
+demographic_wide <- demographic_long |>
+  pivot_wider(
+    names_from = category,
+    values_from = Value
+  )
+
+# Convert 'Ward Number' to numeric in both datasets to allow merging
+total_funding_ward$`Ward Number` <- as.numeric(as.character(total_funding_ward$`Ward Number`))
+demographic_wide$`Ward Number` <- as.numeric(as.character(demographic_wide$`Ward Number`))
+
+# Merge budget data with demographic data
+merged_data <- total_funding_ward |>
+  left_join(demographic_wide, by = "Ward Number")
+
+# Convert 'Total Population' to numeric
+merged_data$`Total Population` <- as.numeric(gsub(",", "", merged_data$`Total Population`))
+
+
+# Create per capita funding variable
+merged_data <- merged_data |>
+  mutate(per_capita_funding = total_funding / `Total Population`)
+
+
+# Convert columns from list to double 
+merged_data$`Bachelor’s degree or higher` <- as.numeric(merged_data$`Bachelor’s degree or higher`)
[…]

### 2024-09-26 scripts/05-analysis.R
Update 05-analysis.R

@@ -259,7 +259,5 @@ ggplot(filtered_data, aes(x = as.numeric(`Median total income of households in 2
   scale_y_continuous(labels = dollar_format()) +
   theme_minimal()
 
-  
-
 # Save merged data to CSV
 write_csv(merged_data, "data/analysis_data/merged_data.csv")

### 2024-09-26 paper/paper.qmd
small formatting changes to yearly funding graph in quarto
Improved abstract

--- 2024-09-25 paper/paper.qmd
@@
 date-format: long
-abstract: "This research examines the allocation of capital funding across Toronto's 25 wards and explores how these allocations correspond with various socioeconomic indicators.
-The analysis reveals significant disparities in per capita funding, with higher-income wards often receiving more funding than lower-income wards."
+abstract: "This research examines the allocation of capital funding across Toronto's 25 wards from 2022 to 2031, examining its relationship with socioeconomic indicators.
+The results reveal significant disparities in per capita funding, with higher-income wards and those with higher educational attainment receiving disproportionately more investment.
+Conversely, wards with higher proportions of low-income households and visible minority populations receive less funding per capita."
 format: pdf
@@
 
-2.  Ward Profiles (2021): \\
-This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
+2.
+Ward Profiles (2021): \\ This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
 It was filtered to remain the variables which were most useful for this research.
 These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population.
-[may need to include others, depends on what I decide to go with in the end.]
+\[may need to include others, depends on what I decide to go with in the end.\]
 
 # Results {#sec-results}
-
-Lots of tables and graphs to be included here once they're properly formatted!
 

### 2024-09-26 paper/paper.qmd
Improved introduction to include other research

@@
 It is also city where its neighbourhoods vary significantly in income, educational attainment, and demographic makeup.
-The allocation of capital funding plays an important role in shaping the infrastructure and services available to residents.
-This impacts their quality of life and access to opportunities.
+The allocation of capital funding plays a crucial role in developing infrastructure and services which directly impact residents' quality of life and access to opportunities.
+Equitable distribution of these funds is essential to ensure that all communities within the city can thrive.
 
-Despite the importance of equitable funding, there is evidence that budget distributions may not align with the needs of different communities.
-\[Write about these two:https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8186789/ , http://neighbourhoodchange.ca/documents/2016/05/walks-etal-2016-income-inequality-toronto-york-rp-238.pdf\]
+Despite the city's commitment to equity, recent studies have raised concerns about the fairness of budget distributions.
+Research by Walks et al.
+(2016) \[REFERENCE THIS\] highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty.
+Similarly, Hulchanski (2010) \[REFERENCE THIS\] discusses the "Three Cities" within Toronto, where socioeconomic divisions are becoming more pronounced.
 
-By analyzing data from the City of Toronto's Capital Budget & Plan, as well as its Ward Profiles, this research finds disparities in per capita funding among the 25 wards.
-The findings indicate that higher-income wards tend to receive more funding per capita than lower-income wards.
+\[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8186789/ , http://neighbourhoodchange.ca/documents/2016/05/walks-etal-2016-income-inequality-toronto-york-rp-238.pdf\]
 
-The remainder of the paper is organized as follows: Section \ref{sec-data} describes the data sources and methodology used in the analysis.
-Section \ref{sec-results} presents the results, highlighting key findings.
-Section \ref{sec-discussion} discussed the implications of the findings as well as limitations and recommendations for next steps.
+By analyzing data from the City of Toronto's Capital Budget & Plan and Ward Profiles, the study uncovers significant disparities in per capita funding.
+Notably, higher-income wards and those with higher educational attainment tend to receive more funding per capita than lower-income wards, suggesting a potential misalignment with the needs of less affluent communities.
+
+The remainder of the research is organized as follows: Section \ref{sec-data} describes the data sources and methodology used in the analysis.
+Section \ref{sec-results} presents the results, highlighting key findings through visualizations and statistical summaries.
+Section \ref{sec-discussion} discusses the implications of these findings, addresses limitations, and offers recommendations for improvements to this research.
 
@@
 
-1.
-Capital Budget & Plan By Ward (2022-2031): \\ This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
+1.  Capital Budget & Plan By Ward (2022-2031):
+
+    This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
 It includes funding allocations for various programs such as Children's Services, Parks, Foresty & Recreation, Transportation Services, and others.
@@
 
-2.
-Ward Profiles (2021): \\ This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
+2.  Ward Profiles (2021):
+
+    \newline
+
+    This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
 It was filtered to remain the variables which were most useful for this research.

### 2024-09-26 paper/paper.qmd
Included summary results table and updated pdf

@@
 subtitle: "An Analysis of the Funding Disparties Across Toronto Wards (2022 - 2031)"
-author: 
-  - Maria Mangru 
+author: Maria Mangru 
 thanks: "Code and data are available at: https://github.com/MariaMangru/Capital-Budget-Allocations-and-Median-Household-Income."
@@
 Research by Walks et al.
-(2016) \[REFERENCE THIS\] highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty.
-Similarly, Hulchanski (2010) \[REFERENCE THIS\] discusses the "Three Cities" within Toronto, where socioeconomic divisions are becoming more pronounced.
+(2016) **\[REFERENCE THIS\]** highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty.
+Similarly, PAPER HERE **\[REFERENCE THIS\]** WHAT IT TALKED ABOUT.
 
@@
 
-    \newline
-
 This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
@@
 These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population.
-\[may need to include others, depends on what I decide to go with in the end.\]
+**\[UPDATE THIS\]**
+
+```{r setting up workspace, echo=FALSE, message=FALSE, warning=FALSE}
+#### Workspace setup ####
+library(dplyr)
+library(tidyr)
+library(readr)
+library(knitr)
+library(kableExtra)
+library(scales)
+library(ggplot2)
+library(stringr)
+library(janitor)
+
+### Data Handling ###
+
+# Read in the cleaned budget data
+budget_data <- read_csv("../data/analysis_data/clean_budget_data.csv")
+
+# Read in the cleaned demographic data
+demographic_data <- read_csv("../data/analysis_data/clean_ward_demo_data.csv")
+
+# Identify year columns in budget data
+year_cols <- as.character(2022:2031)
+
+# Aggregate total funding per ward number
+total_funding_ward <- budget_data |>
+  group_by(`Ward Number`)|>
+  summarise(
+    total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
+  ) |>
+  arrange(desc(total_funding))
+
+# Adjust total funding (amounts are in thousands of dollars)
+total_funding_ward <- total_funding_ward |>
+  mutate(total_funding = total_funding * 1000)
+
+# Pivot demographic data from wide to long format
+demographic_long <- demographic_data |>
+  pivot_longer(
+    cols = -category,
+    names_to = "Ward Number",
+    values_to = "Value"
+  )
+
+# Pivot demographic data back to wide format to get one row per ward
+demographic_wide <- demographic_long |>
+  pivot_wider(
+    names_from = category,
+    values_from = Value
+  )
+
+# Convert 'Ward Number' to numeric in both datasets to allow merging
+total_funding_ward$`Ward Number` <- as.numeric(as.character(total_funding_ward$`Ward Number`))
+demographic_wide$`Ward Number` <- as.numeric(as.character(demographic_wide$`Ward Number`))
+
+# Merge budget data with demographic data
+merged_data <- total_funding_ward |>
+  left_join(demographic_wide, by = "Ward Number")
+
+# Convert 'Total Population' to numeric
+merged_data$`Total Population` <- as.numeric(gsub(",", "", merged_data$`Total Population`))
+
+# Create per capita funding variable
+merged_data <- merged_data |>
+  mutate(per_capita_funding = total_funding / `Total Population`)
+
+# Convert columns from list to double 
+merged_data$`Bachelor’s degree or higher` <- as.numeric(merged_data$`Bachelor’s degree or higher`)
+merged_data$`Total visible minority population` <- as.numeric(merged_data$`Total visible minority population`)
+
+#Create per capita bachelor's degree variable
+merged_data <- merged_data |>
+  mutate(`Bachelor’s degree or higher per capita` = `Bachelor’s degree or higher` / `Total Population`)
+
+#Create per capita visible minority population variable
+merged_data <- merged_data |>
+  mutate(`Total visible minority population per capita` = `Total visible minority population` / `Total Population`)
+
+```
+
+## Data Summary
+
+```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: tbl-demographics-stats
+#| tbl-cap: Capital Funding and Socioeconomic Indicators by Ward
+
+# Select relevant columns for the table
+merged_table_select <- merged_data |>
+  select(
+    `Ward Number`,
+    per_capita_funding,
+    `Median total income of households in 2020 ($)`,
+    `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`,
+    `Bachelor’s degree or higher per capita`,
+    `Total visible minority population per capita`
+  )
+
+# Arrange by the numeric version of per capita funding before formatting
+merged_table_select <- merged_table_select |>
+  arrange(desc(per_capita_funding))
+
+# Format numeric columns with conditional formatting for LaTeX
+merged_table_2 <- merged_table_select |>
+  mutate(
+    # Format Per Capita Funding as dollars
+    per_capita_funding = dollar(per_capita_funding),
+    `Median total income of households in 2020 ($)` = dollar(as.numeric(`Median total income of households in 2020 ($)`)),
+    `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` = paste0(round(as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`) * 100, 2), "%"),
+    `Bachelor’s degree or higher per capita` = paste0(round(as.numeric(`Bachelor’s degree or higher per capita`) * 100, 2), "%"),
+    `Total visible minority population per capita` = paste0(round(as.numeric(`Total visible minority population per capita`) * 100, 2), "%")
+  )
+
+# Create the table with kable and apply styling
+kable(
+  merged_table_2, 
+  col.names = c(
+    "Ward Number", 
+    "Per Capita Funding", 
+    "Household Median Income", 
+    "Low Income Prevalence", 
+    "Bachelor’s Degree or Higher", 
+    "Visible Minority Population"
+  ), 
+  table.attr = 'style="width:100%; word-wrap: break-word;"'
+) |>
+  kable_styling(
+    bootstrap_options = c("condensed"), 
+    full_width = FALSE,  
+    position = "left",  
+    font_size = 7,  
+  ) |>
+  column_spec(1, width = "1cm") |>
+  column_spec(2:6, width = "3cm")
+
+```
 

### 2024-09-26 paper/paper.qmd
began working on the results section and discussion
needs major updates

@@
 
-# Create the table with kable and apply styling
+# Create the table with kable and apply enhanced styling
 kable(
   merged_table_2, 
+  escape = TRUE,                   
   col.names = c(
@@
     "Visible Minority Population"
-  ), 
-  table.attr = 'style="width:100%; word-wrap: break-word;"'
+  )
 ) |>
@@
     position = "left",  
-    font_size = 7,  
+    font_size = 7 
   ) |>
-  column_spec(1, width = "1cm") |>
-  column_spec(2:6, width = "3cm")
+  column_spec(1, width = "2cm") |>
+  column_spec(2:3, width = "3cm") |>
+  column_spec(4, width = "2cm") |>
+  column_spec(5:6, width = "3cm")
 
@@
 # Results {#sec-results}
+
+Exploring the data reveals substantial disparities in per capita capital funding across Toronto's 25 wards.
+The per capita funding ranges dramatically, from as low as 173.22 in Ward 15 to as high as 7,674.47 in Ward 13.
+This significant variation suggests that capital investment is not evenly distributed among the city's wards, prompting questions about the factors influencing these allocations.
+
+## Per Capita Funding and Household Median Income
+
+When comparing per capita funding to household median income, a pattern emerges that suggests a positive relationship between the two variables.
+Wards with higher median household incomes tend to receive more capital funding per capita.
+For instance:
+
+- **Ward 2** has a household median income of **$100,000** and receives **$4,166.43** per capita in funding.
+- **Ward 10**, with a median income of **$89,000**, receives **$4,127.10** per capita.
+- Conversely, **Ward 22**, with a lower median income of **$77,000**, receives only **$192.44** per capita.
+
+**Figure 1** illustrates this relationship, showing a general upward trend.
+While there are exceptions—such as **Ward 13**, which has a median income of **$65,000** but the highest per capita funding—the overall pattern indicates that wealthier wards receive more capital investment.
+
+FIGURE 1 HERE (MAYBE BOTH WITH WARD 13 AND WITHOUT WARD 13)
+
+## Per Capita Funding and Low-Income Prevalence
+
+Analyzing the prevalence of low-income households provides additional insight into funding disparities.
+Wards with higher percentages of low-income households often receive less per capita funding.
+For example:
+
+- **Ward 22** has a low-income prevalence of **15%** and receives **$192.44** per capita.
+- **Ward 5**, with **15%** low-income prevalence, receives **$867.02** per capita.
+- In contrast, **Ward 2** has a low-income prevalence of **8%** and receives **$4,166.43** per capita.
+
+**Figure 2** demonstrates a downward trend, suggesting that as the prevalence of low-income households increases, per capita funding tends to decrease.
+This inverse relationship raises concerns about the equitable distribution of city resources.
+
+INSERT FIGURE 2 HERE
+
+## Per Capita Funding and Educational Attainment
+
+Educational attainment appears to correlate strongly with funding allocations.
+Wards where a higher percentage of residents hold a bachelor's degree or higher tend to receive more per capita funding.
+Notable examples include:
+
+- **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **$4,127.10** per capita.
+- **Ward 11**, with **56.07%** of residents holding a bachelor's degree or higher, receives **$1,637.81** per capita.
+- Conversely, **Ward 7**, with only **14.67%** educational attainment at this level, receives **$852.05** per capita.
+
+**Figure 3** shows a clear upward trend, indicating that higher educational attainment within a ward is associated with increased capital funding per capita.
+
+FIGURE 3 HERE !!!
+
+## Per Capita Funding and Visible Minority Population
+
+The relationship between per capita funding and the percentage of visible minority populations is complex.
+Some wards with higher percentages of visible minorities receive lower per capita funding.
+For instance:
+
+- **Ward 22** has a visible minority population of **82.31%** and receives **$192.44** per capita.
+- **Ward 23**, with a visible minority population of **92.18%**, receives **$1,829.15** per capita.
+- In contrast, **Ward 2**, with a visible minority population of **31.75%**, receives **$4,166.43** per capita.
+
+**Figure 4** suggests a slight negative correlation, where wards with higher visible minority populations may receive less funding per capita.
+However, the relationship is not as pronounced as with other socioeconomic indicators, indicating that additional factors may be at play.
+
+FIGURE 4 HERE!!!
+
+While general trends are observed, outliers like Ward 13 challenge simple interpretations.
+Despite having a lower median household income and higher low-income prevalence, it receives the highest per capita funding.
+This anomaly suggests that specific projects or unique circumstances, such as major infrastructure developments or targeted investments, can significantly impact funding levels in certain wards.
+
+RESULTS NEED A LOT OF FIXING. THERE IS SOME REDUNDENCY. MAYBE A LOT LOL
 

### 2024-09-26 paper/paper.qmd
tried making data points red

@@
     per_capita_funding = dollar(per_capita_funding),
-    `Median total income of households in 2020 ($)` = dollar(as.numeric(`Median total income of households in 2020 ($)`)),
+      `Median total income of households in 2020 ($)` = paste0(
+      dollar(as.numeric(`Median total income of households in 2020 ($)`)),
+      ifelse(as.numeric(`Median total income of households in 2020 ($)`) <= 84000, "**", "")
+    ),
     `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` = paste0(round(as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`) * 100, 2), "%"),
@@
   column_spec(5:6, width = "3cm")
+```
 
-```
+Wards with a **\*\*** indicate a **Household Median Income** of \$84,000 or less.
 
@@
 
-- **Ward 2** has a household median income of **$100,000** and receives **$4,166.43** per capita in funding.
-- **Ward 10**, with a median income of **$89,000**, receives **$4,127.10** per capita.
-- Conversely, **Ward 22**, with a lower median income of **$77,000**, receives only **$192.44** per capita.
+-   **Ward 2** has a household median income of **\$100,000** and receives **\$4,166.43** per capita in funding.
+-   **Ward 10**, with a median income of **\$89,000**, receives **\$4,127.10** per capita.
+-   Conversely, **Ward 22**, with a lower median income of **\$77,000**, receives only **\$192.44** per capita.
 
 **Figure 1** illustrates this relationship, showing a general upward trend.
-While there are exceptions—such as **Ward 13**, which has a median income of **$65,000** but the highest per capita funding—the overall pattern indicates that wealthier wards receive more capital investment.
+While there are exceptions—such as **Ward 13**, which has a median income of **\$65,000** but the highest per capita funding—the overall pattern indicates that wealthier wards receive more capital investment.
 
@@
 
-- **Ward 22** has a low-income prevalence of **15%** and receives **$192.44** per capita.
-- **Ward 5**, with **15%** low-income prevalence, receives **$867.02** per capita.
-- In contrast, **Ward 2** has a low-income prevalence of **8%** and receives **$4,166.43** per capita.
+-   **Ward 22** has a low-income prevalence of **15%** and receives **\$192.44** per capita.
+-   **Ward 5**, with **15%** low-income prevalence, receives **\$867.02** per capita.
+-   In contrast, **Ward 2** has a low-income prevalence of **8%** and receives **\$4,166.43** per capita.
 
@@
 
-- **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **$4,127.10** per capita.
-- **Ward 11**, with **56.07%** of residents holding a bachelor's degree or higher, receives **$1,637.81** per capita.
-- Conversely, **Ward 7**, with only **14.67%** educational attainment at this level, receives **$852.05** per capita.
+-   **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **\$4,127.10** per capita.
+-   **Ward 11**, with **56.07%** of residents holding a bachelor's degree or higher, receives **\$1,637.81** per capita.
+-   Conversely, **Ward 7**, with only **14.67%** educational attainment at this level, receives **\$852.05** per capita.
 
@@
 
-- **Ward 22** has a visible minority population of **82.31%** and receives **$192.44** per capita.
-- **Ward 23**, with a visible minority population of **92.18%**, receives **$1,829.15** per capita.
-- In contrast, **Ward 2**, with a visible minority population of **31.75%**, receives **$4,166.43** per capita.
+-   **Ward 22** has a visible minority population of **82.31%** and receives **\$192.44** per capita.
+-   **Ward 23**, with a visible minority population of **92.18%**, receives **\$1,829.15** per capita.
+-   In contrast, **Ward 2**, with a visible minority population of **31.75%**, receives **\$4,166.43** per capita.
 

### 2024-09-26 paper/paper.qmd
highlighted important rows in the table

@@
     per_capita_funding = dollar(per_capita_funding),
-      `Median total income of households in 2020 ($)` = paste0(
-      dollar(as.numeric(`Median total income of households in 2020 ($)`)),
-      ifelse(as.numeric(`Median total income of households in 2020 ($)`) <= 84000, "**", "")
-    ),
+    `Median total income of households in 2020 ($)` = paste0(dollar(as.numeric(`Median total income of households in 2020 ($)`))),
     `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` = paste0(round(as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`) * 100, 2), "%"),
@@
   ) |>
+  add_footnote("Rows in red indicate a ward where the Household Median Income is $84,000 or less", notation="alphabet") |>
   column_spec(1, width = "2cm") |>
@@
   column_spec(4, width = "2cm") |>
-  column_spec(5:6, width = "3cm")
+  column_spec(5:6, width = "3cm") |>
+  row_spec(1, color = "red") |>
+  row_spec(6, color = "red") |>
+  row_spec(9:10, color = "red") |>
+  row_spec(13:14, color = "red") |>
+  row_spec(17:19, color = "red") |>
+  row_spec(21:24, color = "red")
 ```

### 2024-09-26 paper/paper.qmd
updated quarto to include scatter plots for median income and low income measure

@@
 
-Wards with a **\*\*** indicate a **Household Median Income** of \$84,000 or less.
+Rows in **red** indicate a **Household Median Income** of \$84,000 or less.
 
@@
 
+```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fid-householdincome
+
+# Scatter plot of Per Capita Funding vs Median Household Income with ward 13 removed
+
+filtered_data <- merged_data |>
+  filter(`Ward Number` != 13)
+
+ggplot(filtered_data, aes(x = as.numeric(`Median total income of households in 2020 ($)`), y = per_capita_funding)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(
+    title = "Per Capita Funding vs Median Household Income",
+    x = "Median Household Income ($)",
+    y = "Per Capita Funding ($)"
+  ) +
+  scale_x_continuous(labels = dollar_format()) +
+  scale_y_continuous(labels = dollar_format()) +
+  theme_minimal()
+```
+
 FIGURE 1 HERE (MAYBE BOTH WITH WARD 13 AND WITHOUT WARD 13)
@@
 INSERT FIGURE 2 HERE
+```{r Figure 2, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fid-lowincome
+
+# Scatter plot of Per Capita Funding vs Median Household Income with ward 13 removed
+
+filtered_data <- merged_data |>
+  filter(`Ward Number` != 13)
+
+ggplot(filtered_data, aes(x = as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`), y = per_capita_funding)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(
+    title = "Per Capita Funding vs Prevalence of low income",
+    x = "Prevalence of low income (%)",
+    y = "Per Capita Funding ($)"
+  ) +
+  scale_x_continuous(labels = percent_format()) +
+  scale_y_continuous(labels = dollar_format()) +
+  theme_minimal()
+```
+```
 

### 2024-09-26 paper/paper.qmd
added all 4 scatterplots and a bar graph to quarto

@@
 
+```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-percapitafunding
+# Wards by Per Capita Funding
+ggplot(merged_data, aes(x = reorder(`Ward Number`, per_capita_funding), y = per_capita_funding)) +
+  geom_bar(stat = "identity", fill = "#0072B2") +
+  labs(
+    title = "Wards by Per Capita Funding",
+    x = "Ward Number",
+    y = "Per Capita Funding ($CAD)"
+  ) +
+  scale_y_continuous(labels = dollar_format()) +
+  theme_minimal(base_family = "Helvetica") +
+  theme(
+    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
+    axis.title = element_text(size = 14)
+  )
+```
+
 # Results {#sec-results}
@@
 
-```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fid-householdincome
+```{r Figure 2, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-householdincome
 
@@
 
-FIGURE 1 HERE (MAYBE BOTH WITH WARD 13 AND WITHOUT WARD 13)
-
 ## Per Capita Funding and Low-Income Prevalence
@@
 
-INSERT FIGURE 2 HERE
-```{r Figure 2, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fid-lowincome
+```{r Figure 3, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-lowincome
 
-# Scatter plot of Per Capita Funding vs Median Household Income with ward 13 removed
+# Scatter plot of Per Capita Funding vs Prevalence of low income with ward 13 removed
 
@@
 ```
-```
 
@@
 
-FIGURE 3 HERE !!!
+```{r Figure 4, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-education
+
+# Scatter plot of Per Capita Funding vs Bachelor's Degree of Higher with ward 13 removed
+
+filtered_data <- merged_data |>
+  filter(`Ward Number` != 13)
+
+ggplot(filtered_data, aes(x = as.numeric(`Bachelor’s degree or higher per capita`), y = per_capita_funding)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(
+    title = "Per Capita Funding vs Bachelor's degree or higher",
+    x = "Attained Bachelor's degree or higher (%)",
+    y = "Per Capita Funding ($)"
+  ) +
+  scale_x_continuous(labels = percent_format()) +
+  scale_y_continuous(labels = dollar_format()) +
+  theme_minimal()
+```
 
@@
 
-FIGURE 4 HERE!!!
+```{r Figure 5, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-minority
+
+# Scatter plot of Per Capita Funding vs Minority Population with ward 13 removed
+
+filtered_data <- merged_data |>
+  filter(`Ward Number` != 13)
+
+ggplot(filtered_data, aes(x = as.numeric(`Total visible minority population per capita`), y = per_capita_funding)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(
+    title = "Per Capita Funding vs Visible Minority Population",
+    x = "Visible Minority Population (%)",
+    y = "Per Capita Funding ($)"
+  ) +
+  scale_x_continuous(labels = percent_format()) +
+  scale_y_continuous(labels = dollar_format()) +
+  theme_minimal()
+```
 

### 2024-09-26 paper/paper.qmd
Update abstract and introduction

@@
 date-format: long
-abstract: "This research examines the allocation of capital funding across Toronto's 25 wards from 2022 to 2031, examining its relationship with socioeconomic indicators.
-The results reveal significant disparities in per capita funding, with higher-income wards and those with higher educational attainment receiving disproportionately more investment.
-Conversely, wards with higher proportions of low-income households and visible minority populations receive less funding per capita."
+abstract: "This study examines the allocation of capital funding across Toronto's 25 wards from 2022 to 2031, focusing on its relationship with key socioeconomic factors.
+The results found that wards with higher median household incomes and greater educational attainment receive more captial funding per capita.
+Conversely, wards with higher proportions of low-income households and visible minority populations tend to receive less funding per capita."
 format: pdf
@@
 
-Toronto, often referred to as a cultural mosaic, is home to a diverse population with roots from all around the world.
-It is also city where its neighbourhoods vary significantly in income, educational attainment, and demographic makeup.
-The allocation of capital funding plays a crucial role in developing infrastructure and services which directly impact residents' quality of life and access to opportunities.
-Equitable distribution of these funds is essential to ensure that all communities within the city can thrive.
+Toronto is a city renowned for its cultural diversity, however, it is also a city marked by stark socioeconomic disparities.
+Each year, Toronto allocates funds through its capital budget to support essential services such as public transit, libraries, parks and water facilities.
+[REF: https://www.toronto.ca/wp-content/uploads/2017/11/97f7-A170XXXX_Budget_Basics_Understanding-final-web.pdf].
+This funding shapes the city's infrastructure and directly impacts residents' quality of life and access to opportunities.
 
-Despite the city's commitment to equity, recent studies have raised concerns about the fairness of budget distributions.
-Research by Walks et al.
-(2016) **\[REFERENCE THIS\]** highlights growing income inequality in Toronto's neighborhoods, noting a trend toward the spatial concentration of wealth and poverty.
-Similarly, PAPER HERE **\[REFERENCE THIS\]** WHAT IT TALKED ABOUT.
+Despite Toronto's commitment to equity, evidence suggests that wealthier neighborhoods receive a disproportionate share of public investment.
+Previous reports, such as Walks et al.
+(2016) [REFERENCE HERE], highlight how income segregation leads to affluent areas accumulating more resources, while low-income neighborhoods face underfunded services.
+This pattern enhances socioeconomic disparities and hinders the city's overall progress toward inclusivity.
 
-\[https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8186789/ , http://neighbourhoodchange.ca/documents/2016/05/walks-etal-2016-income-inequality-toronto-york-rp-238.pdf\]
-
-By analyzing data from the City of Toronto's Capital Budget & Plan and Ward Profiles, the study uncovers significant disparities in per capita funding.
-Notably, higher-income wards and those with higher educational attainment tend to receive more funding per capita than lower-income wards, suggesting a potential misalignment with the needs of less affluent communities.
+This research investigates how capital funding from 2022 to 2031 is allocated across Toronto's wards and examines its relationship with median household income, educational attainment, and visible minority populations.
+By identifying patterns in funding distribution, we aim to understand whether current investment strategies align with the needs of all communities.
 
@@
 Section \ref{sec-results} presents the results, highlighting key findings through visualizations and statistical summaries.
-Section \ref{sec-discussion} discusses the implications of these findings, addresses limitations, and offers recommendations for improvements to this research.
+Section \ref{sec-discussion} discusses the implications of these findings, addresses limitations, and offers recommendations for further research.
 
@@
 It was filtered to remain the variables which were most useful for this research.
-These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment, Immigration Status, Visible Minority Population.
-**\[UPDATE THIS\]**
+These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment and Visible Minority Population.
+
+These datasets were merged to relate funding and various socioeconomic factors.
 
@@
 
+This research focuses on ward-level data, filtering out "City Wide" funding, which represents investments distributed across the entire city rather than specific wards.
+
+@tbl-demographics-stats summarizes the results of the research.
+
 ```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
+@fig-percapitafunding shows a bar chart of the per capita funding allocated for each ward.
+
 ```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE}
@@
 Exploring the data reveals substantial disparities in per capita capital funding across Toronto's 25 wards.
-The per capita funding ranges dramatically, from as low as 173.22 in Ward 15 to as high as 7,674.47 in Ward 13.
-This significant variation suggests that capital investment is not evenly distributed among the city's wards, prompting questions about the factors influencing these allocations.
+The per capita funding ranges dramatically, from as low as \$173.22 in Ward 15 to as high as \$7,674.47 in Ward 13.
 
@@
 
-When comparing per capita funding to household median income, a pattern emerges that suggests a positive relationship between the two variables.
-Wards with higher median household incomes tend to receive more capital funding per capita.
-For instance:
+ Across the household incomes, the average value is \$84,000.
+This means that 50% of wards have median household incomes of \$84,000 or less.
+Therefore, we will define a higher income ward as which has a median household income greater than \$84,000.
+Therefore, When comparing per capita funding to household median income, the pattern of wards with higher median household incomes tend to receive more capital funding per capita.
+For instance, of the top 5 wards with the highest per capita funding, 4 were higher income.
+Conversely, of the wards with the 5th lowest per capita funding, 4 were lower income.
 
--   **Ward 2** has a household median income of **\$100,000** and receives **\$4,166.43** per capita in funding.
--   **Ward 10**, with a median income of **\$89,000**, receives **\$4,127.10** per capita.
--   Conversely, **Ward 22**, with a lower median income of **\$77,000**, receives only **\$192.44** per capita.
-
-**Figure 1** illustrates this relationship, showing a general upward trend.
-While there are exceptions—such as **Ward 13**, which has a median income of **\$65,000** but the highest per capita funding—the overall pattern indicates that wealthier wards receive more capital investment.
+**Figure 1** illustrates this relationship.
+Ward 13 was removed from the analysis as it causes the data to be skewed and isn't an accurate representation of the relationship.
+As we can see, the trend line is going upwards which indicates a positive relationship between per capita funding and median household income.
+That is, as median house income increases, per capita funding typically also increases.
 
@@
 Wards with higher percentages of low-income households often receive less per capita funding.
-For example:
-
--   **Ward 22** has a low-income prevalence of **15%** and receives **\$192.44** per capita.
--   **Ward 5**, with **15%** low-income prevalence, receives **\$867.02** per capita.
--   In contrast, **Ward 2** has a low-income prevalence of **8%** and receives **\$4,166.43** per capita.
+A ward is defined as having a higher percentage of low income households if the low income prevalence is 13\% or higher.
+Therefore, of the top 5 wards for per capita funding, two are classified as low income. Conversely, of the bottom 5 wards for per capita funding, four are classified as low income.
 
@@
 
-Educational attainment appears to correlate strongly with funding allocations.
+Educational attainment appears to have a strong positive relationship with funding allocations.
 Wards where a higher percentage of residents hold a bachelor's degree or higher tend to receive more per capita funding.
+A higher percentage of residents holding a bachelor's degree or higher is defined as a 33%.
 Notable examples include:
 
--   **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **\$4,127.10** per capita.
+-   **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **\$4,127.10** per capita, the second highest per capita funding.
 -   **Ward 11**, with **56.07%** of residents holding a bachelor's degree or higher, receives **\$1,637.81** per capita.
@@
 Some wards with higher percentages of visible minorities receive lower per capita funding.
-For instance:
+A higher percentage of visible minorities is defined as 57% or more.
 
--   **Ward 22** has a visible minority population of **82.31%** and receives **\$192.44** per capita.
+Of the top 10 per capita funded wards, 4 have a higher percentage of visible minorities. While for the 10 lowest per capita funded, 6 have a higher percentage of visible minorities. 
+More specifically:
+
+-   **Ward 22** has a visible minority population of **82.31%** and receives **\$192.44** per capita. This is the 2nd lowest per capita funding. 
 -   **Ward 23**, with a visible minority population of **92.18%**, receives **\$1,829.15** per capita.
@@
 
-**Figure 4** suggests a slight negative correlation, where wards with higher visible minority populations may receive less funding per capita.
+**Figure 4** suggests a slight negative relationship, where wards with higher visible minority populations may receive less funding per capita.
 However, the relationship is not as pronounced as with other socioeconomic indicators, indicating that additional factors may be at play.
@@
 
-While general trends are observed, outliers like Ward 13 challenge simple interpretations.
-Despite having a lower median household income and higher low-income prevalence, it receives the highest per capita funding.
-This anomaly suggests that specific projects or unique circumstances, such as major infrastructure developments or targeted investments, can significantly impact funding levels in certain wards.
-
-RESULTS NEED A LOT OF FIXING. THERE IS SOME REDUNDENCY. MAYBE A LOT LOL
+There are outliers such as Ward 23, where 92.18% of residents are visible minorities, yet the ward receives $1,829.15 per capita.
+However, the general trend suggests that wards with larger visible minority populations tend to receive less funding overall, raising concerns about whether capital funding is being equitably distributed in alignment with the needs of these communities.
 

### 2024-09-26 scripts/2024_script.R

@@ -1,0 +1,34 @@
+library(dplyr)
+library(tidyverse)
+library(readr)
+library(stringr)
+
+
+
+
+# Clean budget data 
+budget_data2 <- read_csv("data/raw_data/2024-2033-capital-budget-and-plan-details.csv")
+
+# Remove rows where 'Ward Number' is "CW"
+budget_data <- budget_data2 |>
+  filter(`Ward Number` != "CW")
+
+# Identify year columns
+year_cols <- as.character(2022:2031)
+
+# Function to clean amount 
+clean_funding <- function(x) {
+  x_clean <- ifelse(x == "-", "0", x)
+  x_clean <- str_replace_all(x_clean, ",", "")
+  as.numeric(x_clean)
+}
+
+# Apply the cleaning function to year columns
+budget_data_clean <- budget_data |>
+  mutate(across(all_of(year_cols), clean_funding))
+
+
+# Save cleaned budget data 
+write_csv(budget_data_clean, "data/analysis_data/clean_budget_data2.csv")
+
+

### 2024-09-26 paper/paper.qmd
updated data section to include variables of interest, graphs and short explanations

@@
 Each year, Toronto allocates funds through its capital budget to support essential services such as public transit, libraries, parks and water facilities.
-[REF: https://www.toronto.ca/wp-content/uploads/2017/11/97f7-A170XXXX_Budget_Basics_Understanding-final-web.pdf].
+\[REF: https://www.toronto.ca/wp-content/uploads/2017/11/97f7-A170XXXX_Budget_Basics_Understanding-final-web.pdf\].
 This funding shapes the city's infrastructure and directly impacts residents' quality of life and access to opportunities.
@@
 Previous reports, such as Walks et al.
-(2016) [REFERENCE HERE], highlight how income segregation leads to affluent areas accumulating more resources, while low-income neighborhoods face underfunded services.
+(2016) \[REFERENCE HERE\], highlight how income segregation leads to affluent areas accumulating more resources, while low-income neighborhoods face underfunded services.
 This pattern enhances socioeconomic disparities and hinders the city's overall progress toward inclusivity.
@@
 
-This research utilized two primary datasets provided by the City of Toronto's Open Data Portal.
+This study utilizes two primary datasets provided by the City of Toronto's Open Data Portal.
 
@@
 
-This dataset contains the approved capital budget and plan for each of Toronto's 25 wards over a ten-year period from 2022 to 2031.
-It includes funding allocations for various programs such as Children's Services, Parks, Foresty & Recreation, Transportation Services, and others.
-The data provides a detailed breakdown of how much funding each ward is set to receive for specific projects and services.
+This dataset provides detailed funding allocations for each of Toronto's 25 wards over a ten-year period It covers various programs such as Children's Services, Parks, Forestry & Recreation, and Transportation Services.
 
@@
 
-This dataset provides comprehensive socioeconomic information for each ward based on the 2021 census.
-It was filtered to remain the variables which were most useful for this research.
-These are: Total Population, Median Household Income, Prevalence of Low Income Households, Educational Attainment and Visible Minority Population.
+This dataset provides comprehensive socioeconomic information based on the 2021 census, including total population, median household income, prevalence of low-income households, educational attainment, and visible minority populations.
 
-These datasets were merged to relate funding and various socioeconomic factors.
+These datasets were merged in order to link capital funding allocations with socioeconomic indicators at the ward level.
+The analysis for this paper was conducted using R (REFERENCE), with several packages from the tidyverse suite (REFERENCE WICKHAM ETA) being used for data cleaning, manipulation and visualization.
+Visualizations were created using ggplot2, and tables were rendered using kableExtra.
 
@@
 
-## Data Summary
+## Data Preparation
 
-This research focuses on ward-level data, filtering out "City Wide" funding, which represents investments distributed across the entire city rather than specific wards.
+This research focused on ward-specific funding by excluding "City Wide" allocations that are not tied to particular wards.
+The per capita funding was calculated by dividing each ward's total funding by its population.
+Similarly, the variables `Educational Attainment` and `Visible Minority Population` were were expressed as percentages by dividing each variable by the ward's population.
+This was done to standardized the variables allowing the comparison across wards.
 
-@tbl-demographics-stats summarizes the results of the research.
+The key variables of interest for this research are:
+
+-   Per Capita Funding: Total capital funding allocated to a ward divided by its population.
+
+-   Median Household Income: The median income of households within a ward.
+
+-   Low-Income Prevalence: The percentage of households classified as low-income in a ward.
+
+-   Educational Attainment: The percentage of residents with a bachelor's degree or higher.
+
+-   Visible Minority Population: The percentage of residents who identify as visible minorities.
 
@@
 #| label: tbl-demographics-stats
-#| tbl-cap: Capital Funding and Socioeconomic Indicators by Ward
+#| tbl-cap: Table Showing Capital Funding and Socioeconomic Indicators by Ward
 
@@
 
-@fig-percapitafunding shows a bar chart of the per capita funding allocated for each ward.
+@tbl-demographics-stats shows the ward specific funding and key demographic indicators.
+Ward 13 stands out with the highest per capita funding and a high percentage of low income prevalence, relative to other wards.
+
+\newpage
 
@@
 #| label: fig-percapitafunding
+#| fig-cap: Figure Showing Per Capita Capital Funding by Ward
 # Wards by Per Capita Funding
@@
 ```
+
+@fig-percapitafunding gives a visual representation of the per capita funding allocated to each ward.
+This per capita funding varied widely, ranging from \$173.22 in Ward 15 to \$7,674.47 in Ward 13.
+This significant variation indicates an uneven distribution of resources which may not align with the needs of all communities.
 
@@
 Wards with higher percentages of low-income households often receive less per capita funding.
-A ward is defined as having a higher percentage of low income households if the low income prevalence is 13\% or higher.
-Therefore, of the top 5 wards for per capita funding, two are classified as low income. Conversely, of the bottom 5 wards for per capita funding, four are classified as low income.
+A ward is defined as having a higher percentage of low income households if the low income prevalence is 13% or higher.
+Therefore, of the top 5 wards for per capita funding, two are classified as low income.
+Conversely, of the bottom 5 wards for per capita funding, four are classified as low income.
 
@@
 
-Of the top 10 per capita funded wards, 4 have a higher percentage of visible minorities. While for the 10 lowest per capita funded, 6 have a higher percentage of visible minorities. 
-More specifically:
+Of the top 10 per capita funded wards, 4 have a higher percentage of visible minorities. While for the 10 lowest per capita funded, 6 have a higher percentage of visible minorities. More specifically:
 
@@
 
-There are outliers such as Ward 23, where 92.18% of residents are visible minorities, yet the ward receives $1,829.15 per capita.
+There are outliers such as Ward 23, where 92.18% of residents are visible minorities, yet the ward receives \$1,829.15 per capita.
 However, the general trend suggests that wards with larger visible minority populations tend to receive less funding overall, raising concerns about whether capital funding is being equitably distributed in alignment with the needs of these communities.

### 2024-09-26 paper/paper.qmd
Fixed sentence structure for household median income results

@@
 #| label: fig-percapitafunding
-#| fig-cap: Figure Showing Per Capita Capital Funding by Ward
+#| fig-cap: Box plot showing Per Capita Capital Funding by Ward
+
 # Wards by Per Capita Funding
@@
 
-Exploring the data reveals substantial disparities in per capita capital funding across Toronto's 25 wards.
-The per capita funding ranges dramatically, from as low as \$173.22 in Ward 15 to as high as \$7,674.47 in Ward 13.
-
 ## Per Capita Funding and Household Median Income
 
-Across the household incomes, the average value is \$84,000.
-This means that 50% of wards have median household incomes of \$84,000 or less.
-Therefore, we will define a higher income ward as which has a median household income greater than \$84,000.
-Therefore, When comparing per capita funding to household median income, the pattern of wards with higher median household incomes tend to receive more capital funding per capita.
-For instance, of the top 5 wards with the highest per capita funding, 4 were higher income.
-Conversely, of the wards with the 5th lowest per capita funding, 4 were lower income.
+The average income across wards is \$84,000, meaning half of the wards have median incomes at or below this amount.
+Higher income wards will be categorized as one with a median household income above \$84,000.
+A positive relationship was observed between per capita funding and median household income as wards above the citywide average \$84,000 generally received more funding per capita.
+Specifically, among the top five wards in per capita funding, four have median incomes above \$84,000.
+However, among the bottom five wards, four have median incomes below \$84,000.
 
-**Figure 1** illustrates this relationship.
-Ward 13 was removed from the analysis as it causes the data to be skewed and isn't an accurate representation of the relationship.
-As we can see, the trend line is going upwards which indicates a positive relationship between per capita funding and median household income.
-That is, as median house income increases, per capita funding typically also increases.
+More specifically, Ward 2, with a high median income of \$100,000, received the second-highest per capita funding at \$4,166.43.
+In contrast, Ward 22, which has a lower median income of \$77,000, received the second-lowest per capita funding at just \$192.44.
+This pattern suggests that wealthier wards benefit more from capital investments.
 
@@
 #| label: fig-householdincome
+#| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Median Household Income
 
@@
 ```
+
+@fig-householdincome shows the relationship between the variables, with Ward 13 excluded from the analysis due to its disproportionate impact, which distorts the overall data and does not provide an accurate depiction of the relationship.
+The upward trend line indicates a positive relationship between per capita funding and median household income.
+In other words, as median household income rises, per capita funding tends to increase as well.
 

### 2024-09-27 scripts/05-analysis.R
updated analysis script to include new visual for low income

@@ -259,5 +259,77 @@ ggplot(filtered_data, aes(x = as.numeric(`Median total income of households in 2
   scale_y_continuous(labels = dollar_format()) +
   theme_minimal()
 
+
 # Save merged data to CSV
 write_csv(merged_data, "data/analysis_data/merged_data.csv")
+
+
+
+
+
+
+### Needs cleaning to better integrate with the rest of the script 
+
+
+# Select top 5 and bottom 5 wards by per capita funding 
+top5_wards <- merged_data |>
+  arrange(desc(per_capita_funding)) |>
+  slice(2:6)
+
+bottom5_wards <- merged_data |>
+  arrange(per_capita_funding)|>
+  slice(1:5)
+
+# Combine these wards 
+selected_wards <- bind_rows(top5_wards, bottom5_wards) |>
+  mutate(Category = ifelse(`Ward Number` %in% top5_wards$`Ward Number`, "Top 5", "Bottom 5"))
+
+selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` <- as.numeric(as.character(selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`))
+
+
+# Calculate scaling factor
+max_funding <- max(selected_wards$per_capita_funding, na.rm = TRUE)
+max_low_income <- max(selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`, na.rm = TRUE)
+scale_factor <- max_funding / max_low_income
+
+# Add a scaled Low Income Prevalence for plotting
+selected_wards <- selected_wards |>
+  mutate(Scaled_Low_Income_Prevalence = `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` * scale_factor)
+
+
+# Create the Combined Bar and Line Chart
+ggplot(selected_wards, aes(x = reorder(`Ward Number`, per_capita_funding))) +
+  geom_bar(aes(y = per_capita_funding, fill = Category), stat = "identity") +
+  geom_line(aes(y = Scaled_Low_Income_Prevalence, group = 1, color = "Low Income Prevalence"), size = 1.2) +
+  geom_point(aes(y = Scaled_Low_Income_Prevalence, color = "Low Income Prevalence"), size = 3) +
+ 
+   scale_y_continuous(
+    name = "Per Capita Funding ($CAD)",
+    labels = dollar_format(),
[…]

### 2024-09-27 paper/paper.qmd
updated analysis script to include new visual for low income
changed low income graph to be a bar and line graph, updated results text

@@
 #| label: fig-percapitafunding
-#| fig-cap: Box plot showing Per Capita Capital Funding by Ward
+#| fig-cap: Bar graph showing Per Capita Capital Funding by Ward
 
@@
 
-Analyzing the prevalence of low-income households provides additional insight into funding disparities.
-Wards with higher percentages of low-income households often receive less per capita funding.
-A ward is defined as having a higher percentage of low income households if the low income prevalence is 13% or higher.
-Therefore, of the top 5 wards for per capita funding, two are classified as low income.
-Conversely, of the bottom 5 wards for per capita funding, four are classified as low income.
-
-**Figure 2** demonstrates a downward trend, suggesting that as the prevalence of low-income households increases, per capita funding tends to decrease.
-This inverse relationship raises concerns about the equitable distribution of city resources.
+It was found that an inverse relationship exists between per capita funding and the prevalence of low income households.
+We will classify a ward as having a high percentage of low-income households if the prevalence is 13% or higher, which represents the average low-income prevalence.
+Among the top five funded wards, only two can be classified as higher low-income prevalence.
+In contrast, of the bottom five funded wards, four can be classified as having higher low-income prevalence.
+This suggests that wards with more low-income households receive less capital budget funding.
 
@@
 #| label: fig-lowincome
+#| fig-cap: Per capita funding and low income prevalence for top and bottom 5 wards excluding ward 13
 
-# Scatter plot of Per Capita Funding vs Prevalence of low income with ward 13 removed
+top5_wards <- merged_data |>
+  arrange(desc(per_capita_funding)) |>
+  slice(2:6)
 
-filtered_data <- merged_data |>
-  filter(`Ward Number` != 13)
+bottom5_wards <- merged_data |>
+  arrange(per_capita_funding)|>
+  slice(1:5)
 
-ggplot(filtered_data, aes(x = as.numeric(`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`), y = per_capita_funding)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE) +
+# Combine these wards 
+selected_wards <- bind_rows(top5_wards, bottom5_wards) |>
+  mutate(Category = ifelse(`Ward Number` %in% top5_wards$`Ward Number`, "Top 5", "Bottom 5"))
+
+selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` <- as.numeric(as.character(selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`))
+
+# Calculate scaling factor
+max_funding <- max(selected_wards$per_capita_funding, na.rm = TRUE)
+max_low_income <- max(selected_wards$`Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)`, na.rm = TRUE)
+scale_factor <- max_funding / max_low_income
+
+# Add a scaled Low Income Prevalence for plotting
+selected_wards <- selected_wards |>
+  mutate(Scaled_Low_Income_Prevalence = `Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%)` * scale_factor)
+
+# Create the Combined Bar and Line Chart
+ggplot(selected_wards, aes(x = reorder(`Ward Number`, per_capita_funding))) +
+  geom_bar(aes(y = per_capita_funding, fill = Category), stat = "identity") +
+  geom_line(aes(y = Scaled_Low_Income_Prevalence, group = 1, color = "Low Income Prevalence"), size = 1.2) +
+  geom_point(aes(y = Scaled_Low_Income_Prevalence, color = "Low Income Prevalence"), size = 3) +
+
+   scale_y_continuous(
+    name = "Per Capita Funding ($CAD)",
+    labels = dollar_format(),
+    sec.axis = sec_axis(~./scale_factor, name = "Low Income Prevalence")
+  ) +
   labs(
-    title = "Per Capita Funding vs Prevalence of low income",
-    x = "Prevalence of low income (%)",
-    y = "Per Capita Funding ($)"
+    title = "Top 5 and Bottom 5 Wards by Per Capita Funding with Low Income Prevalence",
+    x = "Ward Number",
+    fill = "Category",
+    color = ""
   ) +
-  scale_x_continuous(labels = percent_format()) +
-  scale_y_continuous(labels = dollar_format()) +
-  theme_minimal()
+
+  scale_fill_manual(values = c("Top 5" = "#0072B2", "Bottom 5" = "red")) +
+  scale_color_manual(values = c("Low Income Prevalence" = "black")) +
+
+  theme_minimal(base_family = "Helvetica") +
+  theme(
+    plot.title = element_text(size = 10, hjust = 0.5),
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
+    axis.title.y.left = element_text(color = "black", size = 10),
+    axis.title.y.right = element_text(color = "black", size = 10),
+    legend.position = "top",
+    legend.title = element_blank(),
+    legend.text = element_text(size = 8)
+  )
 ```
+
+@fig-lowincome compares the top and bottom 5 wards by per capita funding, with the prevalence of low-income households overlaid.
+The red bars represent the wards with the lowest per capita funding, while the blue bars represent those with the highest.
+Ward 13 was excluded from this graph due to its skewed impact on the data.
+The black line shows the prevalence of low-income households, which is generally higher in wards with lower per capita funding.
+However, even among the top-funded wards, the relationship between per capita funding and low-income prevalence is not perfectly linear, suggesting that while funding tends to be higher in wealthier wards, some lower-income wards still receive significant funding.
 

### 2024-09-27 paper/paper.qmd
fixed wording for education results

@@
 
-Educational attainment appears to have a strong positive relationship with funding allocations.
-Wards where a higher percentage of residents hold a bachelor's degree or higher tend to receive more per capita funding.
-A higher percentage of residents holding a bachelor's degree or higher is defined as a 33%.
-Notable examples include:
-
--   **Ward 10**, where **60.97%** of residents have at least a bachelor's degree, receives **\$4,127.10** per capita, the second highest per capita funding.
--   **Ward 11**, with **56.07%** of residents holding a bachelor's degree or higher, receives **\$1,637.81** per capita.
--   Conversely, **Ward 7**, with only **14.67%** educational attainment at this level, receives **\$852.05** per capita.
-
-**Figure 3** shows a clear upward trend, indicating that higher educational attainment within a ward is associated with increased capital funding per capita.
+We will also classify wards where at least 33% of residents hold a bachelor's degree or higher as having high educational attainment.
+There is a clear positive relationship between educational attainment and per capita funding as wards with higher education levels tend to receive more funding.
+More specifically, Ward 10, with 60.97% of residents holding a bachelor’s degree, received \$4,127.10 per capita, while Ward 11, with 56.07%, received \$1,637.81 per capita.
+In contrast, Ward 7, where only 14.67% of residents hold a degree, received just \$852.05 per capita.
 
@@
 #| label: fig-education
+#| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Education Levels
 
@@
 ```
+
+The trend line in @fig-education illustrates that, generally, wards with a higher percentage of degree holders are more likely to see higher per capita funding allocations.
+Ward 13 was excluded from this analysis due to its skewing effect on the data.
 

### 2024-09-27 paper/paper.qmd
fixed sentences for visible minority results

@@
 
-The relationship between per capita funding and the percentage of visible minority populations is complex.
-Some wards with higher percentages of visible minorities receive lower per capita funding.
-A higher percentage of visible minorities is defined as 57% or more.
+The relationship between per capita funding and the proportion of visible minority populations appears complex, with signs of potential disparities.
+Wards with higher visible minority populations often receive less funding per capita.
+We will consider a ward to have a high percentage of visible minorities if 57% or more of its population falls into this category.
+57% represents the average percent of visible minorities across the wards.
+Among the top ten funded wards, only four have higher proportions of visible minorities, while among the bottom ten, six wards have higher visible minority populations.
 
-Of the top 10 per capita funded wards, 4 have a higher percentage of visible minorities. While for the 10 lowest per capita funded, 6 have a higher percentage of visible minorities. More specifically:
-
--   **Ward 22** has a visible minority population of **82.31%** and receives **\$192.44** per capita. This is the 2nd lowest per capita funding.
--   **Ward 23**, with a visible minority population of **92.18%**, receives **\$1,829.15** per capita.
--   In contrast, **Ward 2**, with a visible minority population of **31.75%**, receives **\$4,166.43** per capita.
-
-**Figure 4** suggests a slight negative relationship, where wards with higher visible minority populations may receive less funding per capita.
-However, the relationship is not as pronounced as with other socioeconomic indicators, indicating that additional factors may be at play.
+Examples highlighting this trend include Ward 22, where 82.31% of the population are visible minorities, and which received only \$192.44 per capita, and Ward 23, with 92.18% visible minorities, which received \$1,829.15 per capita.
+In contrast, Ward 2, with a lower visible minority population of 31.75%, received significantly more funding at \$4,166.43 per capita.
+While exceptions exist, the general trend suggests that wards with larger visible minority populations may receive less funding.
 
@@
 #| label: fig-minority
+#| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Minority Population
 
@@
 
-There are outliers such as Ward 23, where 92.18% of residents are visible minorities, yet the ward receives \$1,829.15 per capita.
-However, the general trend suggests that wards with larger visible minority populations tend to receive less funding overall, raising concerns about whether capital funding is being equitably distributed in alignment with the needs of these communities.
+@fig-minority illustrates a downward trend between per capita funding and the percentage of visible minority populations across wards, with Ward 13 excluded.
+As the visible minority population increases, per capita funding tends to decrease.
+This trend suggests potential disparities in how funding is distributed among wards with higher visible minority populations.
 

### 2024-09-27 paper/paper.qmd
wrote a short discussion to give context to results

@@
 These datasets were merged in order to link capital funding allocations with socioeconomic indicators at the ward level.
-The analysis for this paper was conducted using R (REFERENCE), with several packages from the tidyverse suite (REFERENCE WICKHAM ETA) being used for data cleaning, manipulation and visualization.
+The analysis for this paper was conducted using R @citeR, with several packages from the tidyverse suite (REFERENCE WICKHAM ETA) being used for data cleaning, manipulation and visualization.
 Visualizations were created using ggplot2, and tables were rendered using kableExtra.
@@
 
-## First discussion point {#sec-first-point}
+This study's examination of capital funding allocation across Toronto's wards reveals patterns that mirror broader socioeconomic dynamics discussed in urban studies literature.
+The disparities observed align with the findings of Walks et al.
+(2016) [REFERENCE], who highlight how income segregation in Canadian cities leads to affluent areas accumulating more resources while low-income neighborhoods face underfunded services.
+These patterns contribute to socio-spatial income inequality, where wealth and resources are unevenly distributed, reinforcing existing socioeconomic divides.
 
-## Second discussion point
+The unequal distribution of capital funding has significant implications for both present and future generations.
+Underfunded communities often lack access to essential services such as quality education, healthcare, and public transportation.
+The National Academies of Sciences, Engineering, and Medicine (2023) emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
+Limited access to resources in underfunded wards can hinder social mobility and perpetuate cycles of poverty.
 
-## Third discussion point
+Furthermore, the disparities in funding contribute to a widening wealth gap within the city.
+Wealthier wards receiving more investment can improve infrastructure, increase property values, and attract further economic development, as discussed by Walks et al.
+(2016).
+In contrast, underfunded wards may experience deteriorating infrastructure and services, making them less attractive for investment and leading to economic stagnation.
+This dynamic exacerbates socioeconomic disparities and can strain social services, as residents in underfunded areas may rely more heavily on social assistance programs due to a lack of access to resources.
+
+These findings highlight the broader effects of unequal resource distribution on urban communities.
+The reinforcement of existing inequalities through funding allocation can lead to increased social stratification, reduced social cohesion, and limited opportunities for residents in disadvantaged areas.
+Addressing these disparities requires a comprehensive understanding of how funding decisions impact different communities and the implementation of strategies that promote equitable access to resources.
 
 ## Weaknesses and next steps
+
+### Weaknesses
 

### 2024-09-27 paper/paper.qmd
wrote weaknesses and next steps

@@
 Underfunded communities often lack access to essential services such as quality education, healthcare, and public transportation.
-The National Academies of Sciences, Engineering, and Medicine (2023) emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
+The National Academies of Sciences, Engineering, and Medicine (2023) [REFERENCE] emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
 Limited access to resources in underfunded wards can hinder social mobility and perpetuate cycles of poverty.
@@
 Wealthier wards receiving more investment can improve infrastructure, increase property values, and attract further economic development, as discussed by Walks et al.
-(2016).
+(2016) [REFERENCE].
 In contrast, underfunded wards may experience deteriorating infrastructure and services, making them less attractive for investment and leading to economic stagnation.
@@
 ### Weaknesses 
+While this study provides interesting results, there are a few limitations.
+Firstly, the analysis is limited by the focus on median household income, educational attainment, and visible minority populations, potentially missing other important socioeconomic factors like housing costs, employment rates, and age demographics.
+Next, the reliance on 2021 census data and projected capital budgets through 2031 may not account for recent changes.
+Lastly, the methodology also focuses on correlations without examining the underlying causes of funding disparities, such as political priorities or historical patterns.
+
+### Next Steps 
+To address these limitations, future research should expand the range of socioeconomic variables, such as housing costs and employment rates, to provide a more comprehensive understanding of funding distribution.
+Longitudinal studies tracking changes in funding and socioeconomic outcomes would help assess the long-term effects of capital investments.
+Additionally, conducting a comparative study with other cities could reveal best practices for equitable funding.
 

### 2024-09-27 paper/paper.qmd
updated paper.pdf

@@
 These datasets were merged in order to link capital funding allocations with socioeconomic indicators at the ward level.
-The analysis for this paper was conducted using R @citeR, with several packages from the tidyverse suite (REFERENCE WICKHAM ETA) being used for data cleaning, manipulation and visualization.
-Visualizations were created using ggplot2, and tables were rendered using kableExtra.
+The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite being used for data cleaning, manipulation and visualization.
+Visualizations were created using @ggplot2, and tables were rendered using kableExtra.
 

### 2024-09-27 scripts/05 - budget_data_results.R
cleaned up code for analysis.r

@@ -166,6 +166,3 @@ ggplot(funding_yearly_ward, aes(x = as.integer(Year), y = Funding, color = `Ward
     legend.title = element_text(size = 13),
     legend.text = element_text(size = 11)
   )
-
-
-

### 2024-09-27 scripts/05-analysis.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Performs analysis of budget and merged data
+# Purpose: Performs analysis of budget and merged demographic data
 # Author: Maria Mangru
 # Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
 
-#### Workspace setup ####
+#### Workspace Setup ####
 library(dplyr)
 library(tidyr)
 library(readr)
@@ -16,80 +16,62 @@ library(ggplot2)
 library(stringr)
 library(janitor)
 
+#### Data Handling ####
 
-### Data Handling ###
+# Define file paths
+budget_file <- "data/analysis_data/clean_budget_data.csv"
+demo_file <- "data/analysis_data/clean_ward_demo_data.csv"
 
-# Read in the cleaned budget data
-budget_data <- read_csv("data/analysis_data/clean_budget_data.csv")
-
-# Read in the cleaned demographic data
-demographic_data <- read_csv("data/analysis_data/clean_ward_demo_data.csv")
+# Read in the cleaned budget and demographic data
+budget_data <- read_csv(budget_file)
+demographic_data <- read_csv(demo_file)
 
 # Identify year columns in budget data
 year_cols <- as.character(2022:2031)
 
-# Aggregate total funding per ward number
+# Aggregate total funding per ward
 total_funding_ward <- budget_data |>
-  group_by(`Ward Number`)|>
+  group_by(`Ward Number`) |>
   summarise(
     total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
   ) |>
-  arrange(desc(total_funding))
+  arrange(desc(total_funding)) |>
+  mutate(total_funding = total_funding * 1000)  # Adjust funding to dollars
 
-# Adjust total funding (amounts are in thousands of dollars)
-total_funding_ward <- total_funding_ward |>
-  mutate(total_funding = total_funding * 1000)
-
-# Pivot demographic data from wide to long format
-demographic_long <- demographic_data |>
+# Pivot demographic data from wide to long, then back to wide to ensure one row per ward
+demographic_wide <- demographic_data |>
   pivot_longer(
     cols = -category,
     names_to = "Ward Number",
     values_to = "Value"
-  )
-
-# Pivot demographic data back to wide format to get one row per ward
-demographic_wide <- demographic_long |>
+  ) |>
   pivot_wider(
     names_from = category,
     values_from = Value
   )
 
-# Convert 'Ward Number' to numeric in both datasets to allow merging
[…]

### 2024-09-27 scripts/2024_script.R

@@ -1,34 +1,0 @@
-library(dplyr)
-library(tidyverse)
-library(readr)
-library(stringr)
-
-
-
-
-# Clean budget data 
-budget_data2 <- read_csv("data/raw_data/2024-2033-capital-budget-and-plan-details.csv")
-
-# Remove rows where 'Ward Number' is "CW"
-budget_data <- budget_data2 |>
-  filter(`Ward Number` != "CW")
-
-# Identify year columns
-year_cols <- as.character(2022:2031)
-
-# Function to clean amount 
-clean_funding <- function(x) {
-  x_clean <- ifelse(x == "-", "0", x)
-  x_clean <- str_replace_all(x_clean, ",", "")
-  as.numeric(x_clean)
-}
-
-# Apply the cleaning function to year columns
-budget_data_clean <- budget_data |>
-  mutate(across(all_of(year_cols), clean_funding))
-
-
-# Save cleaned budget data 
-write_csv(budget_data_clean, "data/analysis_data/clean_budget_data2.csv")
-
-

### 2024-09-27 scripts/05 - budget_data_results.R
deleted unneeded files

@@ -1,168 +1,0 @@
-#### Preamble ####
-# Purpose: Explores data in an attempt to understand it more
-# Author: Maria Mangru
-# Date: September 2024
-# Contact: maria.mangru@mail.utoronto.ca
-# License: MIT
-
-#### Workspace setup ####
-library(dplyr)
-library(tidyverse)
-library(readr)
-library(knitr)
-library(kableExtra)
-library(scales)
-
-### Data handling ###
-# Import cleaned data 
-budget_data_clean <- read_csv("data/analysis_data/2022-2031-budget_data.csv")
-
-# Identify year columns
-year_cols <- as.character(2022:2031)
-
-
-# Reshape data to long format
-budget_data_long <- budget_data_clean |>
-  pivot_longer(
-    cols = all_of(year_cols),
-    names_to = "Year",
-    values_to = "Funding"
-  )
-
-
-### Summing funding values ### 
-
-# Total overall funding per ward number 
-total_funding_ward <- budget_data_clean |>
-  group_by(`Ward Number`) |>
-  summarise(
-    Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)
-  ) |>
-  arrange(desc(Total_Funding))
-
-
-# Funding for each year per ward number 
-funding_yearly_ward <- budget_data_long |>
-  group_by(`Ward Number`, Year) |>
-  summarise(Funding = sum(Funding, na.rm = TRUE)) |>
-  arrange(`Ward Number`, Year)
-
-
-# Total funding per program/agency per ward number 
-total_funding_program_ward <- budget_data_clean |>
-  group_by(`Ward Number`, `Program/Agency Name`) |>
-  summarise(Total_Funding = sum(across(all_of(year_cols)), na.rm = TRUE)) |>
-  arrange(`Ward Number`, desc(Total_Funding))
-
-
-# Yearly funding per program/agency per ward number - maybe remove later
-funding_yearly_program_ward <- budget_data_long |>
-  group_by(`Ward Number`, `Program/Agency Name`, Year) |>
-  summarise(
-    Funding = sum(Funding, na.rm = TRUE)
-  ) |>
-  arrange(`Ward Number`, `Program/Agency Name`, Year)
-
-
-### Tables and Visualizations ###
-
-## Tables ## 
-
-# Total Overall Funding per Ward Number
-total_funding_ward_table <- total_funding_ward |>
-  arrange(desc(Total_Funding)) |>
[…]

### 2024-09-27 scripts/merge_test.R

@@ -1,109 +1,0 @@
-#### Preamble ####
-# Purpose: Merge demographic and budget data 
-# Author: Maria Mangru
-# Date: September 2024
-# Contact: maria.mangru@mail.utoronto.ca
-# License: MIT
-
-
-
-
-#### Workspace Setup ####
-library(dplyr)
-library(tidyr)
-library(readr)
-library(knitr)
-library(kableExtra)
-library(scales)
-library(stringr)
-library(janitor)
-library(ggplot2)
-
-### Data Handling ###
-
-# Read in the budget data
-budget_data <- read_csv("data/analysis_data/20221-2031-budget_data.csv")
-
-# Read in the demographic data
-demographic_data <- read_csv("data/analysis_data/2023-WardProfile-Data.csv")
-
-# View the structure of the datasets
-str(budget_data)
-str(demographic_data)
-
-
-# Identify year columns in budget data
-year_cols <- as.character(2022:2031)
-
-
-# Aggregate total funding per ward number
-total_funding_ward <- budget_data |>
-  group_by(`Ward Number`) |>
-  summarise(
-    total_funding = sum(across(all_of(year_cols)), na.rm = TRUE)
-  ) |>
-  arrange(desc(total_funding))
-
-# View total funding per ward
-print(total_funding_ward)
-
-
-# Pivot demographic data from wide to long format
-demographic_long <- demographic_data |>
-  pivot_longer(
-    cols = -Category,
-    names_to = "Ward Number",
-    values_to = "Value"
-  )
-
-
-# Pivot demographic data back to wide format to get one row per ward
-demographic_wide <- demographic_long |>
-  pivot_wider(
-    names_from = Category,
-    values_from = Value
-  )
-
-demographic_wide$`Ward Number` <- as.numeric(as.character(demographic_wide$`Ward Number`))
-
-
-as.numeric(as.character(total_funding_ward$`Ward Number`))
-
-sapply(total_funding_ward, class)
-sapply(demographic_wide, class)
-
-# Merge budget data with demographic data
-merged_data <- total_funding_ward |>
-  left_join(demographic_wide, by = "Ward Number")
-
-
-colnames(merged_data)
-
-
-
-### Tables and Visualizations ###
-
-# Select relevant columns for the table
-merged_table <- merged_data |>
-  select(
-    `Ward Number`,
-    `total_funding`,
-    `Total Population`,
-    `Median total income of households in 2020 ($)`,
[…]

### 2024-09-27 scripts/04 - test_data.R
updated tests

@@ -7,98 +7,84 @@
 
 
 #### Workspace setup ####
-library(tidyverse)   
-library(testthat)     
-library(kableExtra)  
-library(scales) 
+library(tidyverse)
+library(testthat)
+library(kableExtra)
+library(scales)
 library(readr)
+library(here)
 
-#### Load Data ####
 
-# Cleaned ward data
-ward_data_clean <- read_csv("../data/analysis_data/2023-WardProfile-Data.csv")
+#### Load Cleaned Data ####
+# Load cleaned budget data and ward profile data
+budget_data_clean <- read_csv(here("data", "analysis_data", "clean_budget_data.csv"))
+ward_data_clean <- read_csv(here("data", "analysis_data", "clean_ward_demo_data.csv"))
 
-# Cleaned budget data
-budget_data_clean <- read_csv("../data/analysis_data/2022-2031-budget_data.csv")
-
-# Define year columns
+# Identify year columns
 year_cols <- as.character(2022:2031)
 
-#### Tests ####
 
+### Tests ###
+
+# Test that funding amounts are numeric and correctly handled 
 test_that("Funding Amounts are Numeric and Correctly Handled", {
-  expect_true(all(sapply(ward_data_clean[year_cols], is.numeric)))
-  expect_true(all(is.na(ward_data_clean[year_cols]) | ward_data_clean[year_cols] >= -Inf))
-})
-
-
-test_that("Negative Funding Amounts are Preserved", {
-  negative_funding <- ward_data_clean |>
-    filter(`Funding` < 0)
+  # Check that all year columns are numeric
+  expect_true(
+    budget_data_clean %>%
+      select(all_of(year_cols)) %>%
+      summarise_all(is.numeric) %>%
+      all()
+  )
   
-  expect_true(nrow(negative_funding) > 0)
-})
-
-test_that("Ward Numbers are Consistently Formatted", {
-  expect_true(all(str_detect(ward_data_clean$`Ward Number`, "^Ward \\d+$")))
-})
-
-test_that("Program/Agency Names are Consistently Formatted", {
-  expect_true(all(str_length(ward_data_clean$`Program/Agency Name`) > 0))
+  expect_true(
+    budget_data_clean %>%
+      select(all_of(year_cols)) %>%
+      mutate_all(is.finite) %>%
+      all()
+  )
 })
 
 
-# Total Overall Funding per Ward Number
-total_funding_ward_test <- ward_data_clean |>
-  group_by(`Ward Number`) |>
-  summarise(
[…]

### 2024-09-27 scripts/02 - download_data.R
deleted files not needed

@@ -22,7 +22,7 @@ raw_budget_data <- show_package("7d3bcf2f-8eca-4ed5-a352-a34adb130931")
 raw_budget_data
 
 #### Save data ####
-write_csv(ward_profile, "data/raw_data/WardData.csv") 
-write_csv(ward_profile, "data/raw_data/BudgetData.csv") 
+write_csv(ward_profile, "data/raw_data/2023-WardProfiles-2011-2021-CensusData.csv") 
+write_csv(ward_profile, "data/raw_data/2022-2031-capital-budget-and-plan-details.csv") 
 
          

### 2024-09-27 paper/paper.qmd
updated llm usage and changed paper title

@@
 ---
-title: "Capital Budget Funding in Toronto"
-subtitle: "An Analysis of the Funding Disparties Across Toronto Wards (2022 - 2031)"
+title: "Uneven Investments in Toronto's Wards"
+subtitle: "An Analysis of Capital Funding Allocation in Toronto's Wards from 2022 to 2031"
 author: Maria Mangru 

### 2024-09-27 scripts/05 - analysis.R

@@ -272,5 +272,3 @@ ggplot(selected_wards, aes(x = reorder(`Ward Number`, per_capita_funding))) +
     legend.title = element_blank(),
     legend.text = element_text(size = 12)
   )
-
-

### 2024-09-27 paper/paper.qmd
Update summary table to include ward number and name

@@
 
+# Create a data frame with Ward Names 
+ward_names_df <- data.frame(
+  `Ward Number` = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 
+                    11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 
+                    21, 22, 23, 24, 25),
+  `Ward Name` = c("Etobicoke North", "Etobicoke Centre", "Etobicoke—Lakeshore", "Parkdale—High Park", "York South—Weston", "York Centre",
+                "Humber River—Black Creek", "Eglinton—Lawrence", "Davenport", "Spadina—Fort York", "University—Rosedale", "Toronto—St.
+Paul's", "Toronto Centre", "Toronto—Danforth", "Don Valley West", "Don Valley East", "Don Valley North", "Willowdale", "Beaches—East York", "Scarborough Southwest", "Scarborough Centre", "Scarborough—Agincourt", "Scarborough North", "Scarborough Guildwood", "Scarborough—Rouge Park"),
+  check.names = FALSE
+)
+
+# Join ward names into marged data
+merged_data <- merged_data |>
+  left_join(ward_names_df, by = "Ward Number")
+
 ```
@@
 
+# Create a combined Ward column
+merged_data <- merged_data |>
+  mutate(`Ward` = paste(`Ward Name`, "-", `Ward Number`))
+
 # Select relevant columns for the table
@@
   select(
-    `Ward Number`,
+    `Ward`, 
     per_capita_funding,
@@
   col.names = c(
-    "Ward Number", 
+    "Ward Name - Number", 
     "Per Capita Funding", 

### 2024-09-27 paper/paper.qmd
updated summary table and per capita ward graph
updated summary table to include ward name and footnote, update per capita ward graph to include ward name as well as make more readable

@@
   ) |>
-  add_footnote("Rows in red indicate a ward where the Household Median Income is $84,000 or less", notation="alphabet") |>
-  column_spec(1, width = "2cm") |>
-  column_spec(2:3, width = "3cm") |>
+  footnote(general =  "Rows in red indicate a ward where the Household Median Income is $84,000 or less.") |>
+  column_spec(1, width = "4cm") |>
+  column_spec(2, width = "2cm") |>
+  column_spec(3, width = "3cm") |>
   column_spec(4, width = "2cm") |>
@@
 # Wards by Per Capita Funding
-ggplot(merged_data, aes(x = reorder(`Ward Number`, per_capita_funding), y = per_capita_funding)) +
+ggplot(merged_data, aes(x = reorder(`Ward`, per_capita_funding), y = per_capita_funding)) +
   geom_bar(stat = "identity", fill = "#0072B2") +
@@
     title = "Wards by Per Capita Funding",
-    x = "Ward Number",
+    x = "Ward Name - Number",
     y = "Per Capita Funding ($CAD)"
@@
     plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
-    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
-    axis.title = element_text(size = 14)
-  )
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
+    axis.title = element_text(size = 14),
+    axis.text.y = element_text(size = 10)
+  ) +
+  coord_flip()
 ```
@@
 The disparities observed align with the findings of Walks et al.
-(2016) [REFERENCE], who highlight how income segregation in Canadian cities leads to affluent areas accumulating more resources while low-income neighborhoods face underfunded services.
+(2016) \[REFERENCE\], who highlight how income segregation in Canadian cities leads to affluent areas accumulating more resources while low-income neighborhoods face underfunded services.
 These patterns contribute to socio-spatial income inequality, where wealth and resources are unevenly distributed, reinforcing existing socioeconomic divides.
@@
 Underfunded communities often lack access to essential services such as quality education, healthcare, and public transportation.
-The National Academies of Sciences, Engineering, and Medicine (2023) [REFERENCE] emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
+The National Academies of Sciences, Engineering, and Medicine (2023) \[REFERENCE\] emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
 Limited access to resources in underfunded wards can hinder social mobility and perpetuate cycles of poverty.
@@
 Wealthier wards receiving more investment can improve infrastructure, increase property values, and attract further economic development, as discussed by Walks et al.
-(2016) [REFERENCE].
+(2016) \[REFERENCE\].
 In contrast, underfunded wards may experience deteriorating infrastructure and services, making them less attractive for investment and leading to economic stagnation.
@@
 ### Weaknesses
+
 While this study provides interesting results, there are a few limitations.
@@
 ### Next Steps
+
 To address these limitations, future research should expand the range of socioeconomic variables, such as housing costs and employment rates, to provide a more comprehensive understanding of funding distribution.

### 2024-09-27 paper/paper.qmd
updated low income prevalence graph

@@
   ) |>
-  footnote(general =  "Rows in red indicate a ward where the Household Median Income is $84,000 or less.") |>
+  footnote(general =  "Rows in red indicate wards with a Household Median Income at or below the average of $84,000.") |>
   column_spec(1, width = "4cm") |>
@@
 ```
-
-Rows in **red** indicate a **Household Median Income** of \$84,000 or less.
 
@@
   scale_y_continuous(labels = dollar_format()) +
-  theme_minimal(base_family = "Helvetica") +
+  theme_minimal() +
   theme(
-    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
+    plot.title = element_text(size = 16, hjust = 0.5),
     axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
@@
 selected_wards <- bind_rows(top5_wards, bottom5_wards) |>
-  mutate(Category = ifelse(`Ward Number` %in% top5_wards$`Ward Number`, "Top 5", "Bottom 5"))
+  mutate(Category = ifelse(`Ward` %in% top5_wards$`Ward`, "Top 5", "Bottom 5"))
 
@@
 # Create the Combined Bar and Line Chart
-ggplot(selected_wards, aes(x = reorder(`Ward Number`, per_capita_funding))) +
+ggplot(selected_wards, aes(x = reorder(`Ward`, per_capita_funding))) +
   geom_bar(aes(y = per_capita_funding, fill = Category), stat = "identity") +
@@
     labels = dollar_format(),
-    sec.axis = sec_axis(~./scale_factor, name = "Low Income Prevalence")
+    sec.axis = sec_axis(~./scale_factor, name = "Low Income Prevalence (Decimal)")
   ) +
@@
     title = "Top 5 and Bottom 5 Wards by Per Capita Funding with Low Income Prevalence",
-    x = "Ward Number",
+    x = "Ward Name - Number",
     fill = "Category",
@@
 
-  theme_minimal(base_family = "Helvetica") +
+  theme_minimal() +
   theme(

### 2024-09-27 paper/paper.qmd
updated weakness and next steps

@@
 
-While this study provides interesting results, there are a few limitations.
+While this study provides interesting results, there are a some limitations.
 Firstly, the analysis is limited by the focus on median household income, educational attainment, and visible minority populations, potentially missing other important socioeconomic factors like housing costs, employment rates, and age demographics.
-Next, the reliance on 2021 census data and projected capital budgets through 2031 may not account for recent changes.
-Lastly, the methodology also focuses on correlations without examining the underlying causes of funding disparities, such as political priorities or historical patterns.
+Additionally, the reliance on 2021 census data and projected capital budgets through 2031 may not account for recent changes.
+Lastly, the methodology highlights associations without examining the underlying causes of funding disparities, such as political priorities or historical patterns.
 
@@
 
-To address these limitations, future research should expand the range of socioeconomic variables, such as housing costs and employment rates, to provide a more comprehensive understanding of funding distribution.
-Longitudinal studies tracking changes in funding and socioeconomic outcomes would help assess the long-term effects of capital investments.
-Additionally, conducting a comparative study with other cities could reveal best practices for equitable funding.
+To address these limitations, future research should incorporate a broader range of socioeconomic variables, such as housing costs and employment rates, to gain a more comprehensive understanding of funding distribution.
+Comparative analyses with other cities could also help identify best practices for equitable funding and inform more effective policy decisions
 

### 2024-09-27 paper/paper.qmd
included data meausrement

@@
 
-This study utilizes two primary datasets provided by the City of Toronto's Open Data Portal.
+This study utilizes two primary datasets provided by the City of Toronto's Open Data Portal [REFERENCE].
 
@@
 This dataset provides detailed funding allocations for each of Toronto's 25 wards over a ten-year period It covers various programs such as Children's Services, Parks, Forestry & Recreation, and Transportation Services.
+The budget amounts are in thousands of dollars where a value of 10 represents \$10,000.
+Additionally, the dataset includes both immediate five-year budgets (Year 1 to Year 5) and subsequent five-year planned budgets (Year 6 to Year 10), with a cumulative total for the ten-year span.
 
@@
 
-This dataset provides comprehensive socioeconomic information based on the 2021 census, including total population, median household income, prevalence of low-income households, educational attainment, and visible minority populations.
+Based on the 25-Ward model effective from December 1, 2018, this dataset provides comprehensive socioeconomic information derived from the 2021 Census.
+It includes metrics such as total population, median household income, prevalence of low-income households, educational attainment, and visible minority populations.
 
@@
 The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite being used for data cleaning, manipulation and visualization.
-Visualizations were created using @ggplot2, and tables were rendered using kableExtra.
+Visualizations were created using @ggplot2, and tables were rendered using kableExtra [REFERENCE].
+
+### Data Collection and Measurement
+
+The Capital Budget & Plan By Ward dataset is manually updated and published annually by the City of Toronto, capturing both approved and planned funding across various programs and projects within each ward.
+Key attributes include program or agency names, project and sub-project names, annual budget allocations from Year 1 to Year 10, total ten-year budgets, ward numbers and names, and project categories such as infrastructure or community services.
+
+The Ward Profiles dataset includes a wide range of demographic, social, and economic indicators such as population, household information, migration and mobility patterns, indigenous and ethnocultural composition, language and education levels, as well as labor force and income statistics.
+It should be acknowledged that census data may be subject to sampling and non-response biases, though adjustments are made to mitigate these issues.
+
+Both datasets present certain limitations.
+The Capital Budget & Plan By Ward dataset’s manual updates introduce potential for human error and delays in reflecting the most current budget allocations.
+Similarly, the Ward Profiles dataset excludes data from the 2011 National Household Survey, limiting longitudinal analyses beyond a decade.
+Adjustments made to align older Census data with current ward boundaries may also introduce inaccuracies or inconsistencies.
 
@@
 Similarly, the variables `Educational Attainment` and `Visible Minority Population` were were expressed as percentages by dividing each variable by the ward's population.
-This was done to standardized the variables allowing the comparison across wards.
+This was done to standardized the variables allowing the comparison across wards, ensuring that variations in population sizes do not skew the analysis.
 
@@
 -   Visible Minority Population: The percentage of residents who identify as visible minorities.
+
+Although multiple ten-year budget datasets are available, including the 2023-2032 and 2024-2033 Capital Budget & Plan, this study opted to utilize the 2022-2031 Capital Budget & Plan.
+This is because, the 2022-2031 budget period aligns more closely with the 2021 Ward Profiles, ensuring that the socioeconomic variables and funding allocations are analyzed within a similar timeframe.
 
@@
 To address these limitations, future research should incorporate a broader range of socioeconomic variables, such as housing costs and employment rates, to gain a more comprehensive understanding of funding distribution.
-Comparative analyses with other cities could also help identify best practices for equitable funding and inform more effective policy decisions
+Comparative analyses with other cities could also help identify best practices for equitable funding and inform more effective policy decisions.
+Lastly, conducting a longitudinal study of multiple capital budgets over time would be beneficial in evaluating long-term trends in capital investments.
 

### 2024-09-27 paper/paper.qmd
update references and dates

@@
 Toronto is a city renowned for its cultural diversity, however, it is also a city marked by stark socioeconomic disparities.
-Each year, Toronto allocates funds through its capital budget to support essential services such as public transit, libraries, parks and water facilities.
-\[REF: https://www.toronto.ca/wp-content/uploads/2017/11/97f7-A170XXXX_Budget_Basics_Understanding-final-web.pdf\].
+Each year, Toronto allocates funds through its capital budget to support essential services such as public transit, libraries, parks and water facilities as outlined in @toronto_budget.
 This funding shapes the city's infrastructure and directly impacts residents' quality of life and access to opportunities.
@@
 Despite Toronto's commitment to equity, evidence suggests that wealthier neighborhoods receive a disproportionate share of public investment.
-Previous reports, such as Walks et al.
-(2016) \[REFERENCE HERE\], highlight how income segregation leads to affluent areas accumulating more resources, while low-income neighborhoods face underfunded services.
+Previous reports, such as @walks_report, highlight how income segregation leads to affluent areas accumulating more resources, while low-income neighborhoods face underfunded services.
 This pattern enhances socioeconomic disparities and hinders the city's overall progress toward inclusivity.
@@
 
-This study utilizes two primary datasets provided by the City of Toronto's Open Data Portal [REFERENCE].
+This study utilizes two primary datasets provided by the City of Toronto's @open_data_toronto.
 
@@
 These datasets were merged in order to link capital funding allocations with socioeconomic indicators at the ward level.
-The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite being used for data cleaning, manipulation and visualization.
-Visualizations were created using @ggplot2, and tables were rendered using kableExtra [REFERENCE].
+The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite such as @readr which was used to read in files, @stringr which was used for data cleaning and @dplyr which was used for data manipulation.
+Visualizations were created using @ggplot2, and tables were rendered using @knitr.
 
@@
 This study's examination of capital funding allocation across Toronto's wards reveals patterns that mirror broader socioeconomic dynamics discussed in urban studies literature.
-The disparities observed align with the findings of Walks et al.
-(2016) \[REFERENCE\], who highlight how income segregation in Canadian cities leads to affluent areas accumulating more resources while low-income neighborhoods face underfunded services.
+The disparities observed align with the findings of @walks_report, who highlight how income segregation in Canadian cities leads to affluent areas accumulating more resources while low-income neighborhoods face underfunded services.
 These patterns contribute to socio-spatial income inequality, where wealth and resources are unevenly distributed, reinforcing existing socioeconomic divides.
@@
 Underfunded communities often lack access to essential services such as quality education, healthcare, and public transportation.
-The National Academies of Sciences, Engineering, and Medicine (2023) \[REFERENCE\] emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
+The artcile @poverty_report emphasize that children growing up in low-income households are more likely to experience intergenerational poverty, which can persist without targeted interventions.
 Limited access to resources in underfunded wards can hinder social mobility and perpetuate cycles of poverty.
@@
 Furthermore, the disparities in funding contribute to a widening wealth gap within the city.
-Wealthier wards receiving more investment can improve infrastructure, increase property values, and attract further economic development, as discussed by Walks et al.
-(2016) \[REFERENCE\].
+Wealthier wards receiving more investment can improve infrastructure, increase property values, and attract further economic development, as discussed by @walks_report.
 In contrast, underfunded wards may experience deteriorating infrastructure and services, making them less attractive for investment and leading to economic stagnation.

### 2024-09-27 scripts/01 - simulate_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose:Simulate data 
 # Author: Maria Mangru
-# Date: September 2024
+# Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-09-27 scripts/02 - download_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Downloads data from Open Data Toronto 
 # Author: Maria Mangru
-# Date: September 2024
+# Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-09-27 scripts/04 - test_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests the analysis datasets created 
 # Author: Maria Mangru
-# Date: September 2024
+# Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-09-27 paper/paper.qmd
finished references

@@
 These datasets were merged in order to link capital funding allocations with socioeconomic indicators at the ward level.
-The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite such as @readr which was used to read in files, @stringr which was used for data cleaning and @dplyr which was used for data manipulation.
+The analysis for this paper was conducted using @citeR, with several packages from the @tidyverse suite such as @readr (for reading in files), @stringr (for data cleaning) and @dplyr (for data manipulation).
 Visualizations were created using @ggplot2, and tables were rendered using @knitr.

### 2024-09-27 paper/paper.qmd
update a few lines in abstract and intro

@@
 The results found that wards with higher median household incomes and greater educational attainment receive more captial funding per capita.
-Conversely, wards with higher proportions of low-income households and visible minority populations tend to receive less funding per capita."
+Conversely, wards with higher proportions of low-income households and visible minority populations tend to receive less funding per capita.
+These disparities emphasize the need for a more balanced approach to resource distribution to ensure all communities receive fair and adequate support"
 format: pdf
@@
 By identifying patterns in funding distribution, we aim to understand whether current investment strategies align with the needs of all communities.
+The findings reveal that wards with higher incomes and educational levels receive more funding per capita, while those with more low-income and higher minority population receive less.
 

### 2024-09-27 scripts/01 - simulate_data.R
renamed ward demographic csv

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose:Simulate data 
+# Purpose:Simulate ward profile and budget data
 # Author: Maria Mangru
 # Date: 27 September 2024
 # Contact: maria.mangru@mail.utoronto.ca
@@ -14,7 +14,7 @@ library(readr)
 
 # Simulate ward profile data 
 simulate_ward_profile <- function() {
-  set.seed(403)
+  set.seed(403) # for reproducibility 
   
   ward_numbers <- paste0("Ward", 1:25)
   census_years <- 2011:2021

### 2024-09-27 scripts/03 - data_cleaning.R

@@ -23,7 +23,7 @@ ward_profile <- read_csv("data/raw_data/2023-WardProfiles-2011-2021-CensusData.c
 ward_profile <- na.omit(ward_profile)
 
 # Save cleaned ward profile data 
-write_csv(ward_profile, "data/analysis_data/clean_ward_demo_data.csv")
+write_csv(ward_profile, "data/analysis_data/clean_ward_data.csv")
 
 
 

### 2024-09-27 scripts/04 - test_data.R

@@ -18,7 +18,7 @@ library(here)
 #### Load Cleaned Data ####
 # Load cleaned budget data and ward profile data
 budget_data_clean <- read_csv(here("data", "analysis_data", "clean_budget_data.csv"))
-ward_data_clean <- read_csv(here("data", "analysis_data", "clean_ward_demo_data.csv"))
+ward_data_clean <- read_csv(here("data", "analysis_data", "clean_ward_data.csv"))
 
 # Identify year columns
 year_cols <- as.character(2022:2031)

### 2024-09-27 scripts/05 - analysis.R

@@ -20,7 +20,7 @@ library(janitor)
 
 # Define file paths
 budget_file <- "data/analysis_data/clean_budget_data.csv"
-demo_file <- "data/analysis_data/clean_ward_demo_data.csv"
+demo_file <- "data/analysis_data/clean_ward_data.csv"
 
 # Read in the cleaned budget and demographic data
 budget_data <- read_csv(budget_file)

### 2024-09-27 paper/paper.qmd
renamed ward demographic csv
Included a map of Toronto

@@
 Section \ref{sec-discussion} discusses the implications of these findings, addresses limitations, and offers recommendations for further research.
+The map provided below was produced by @toronto_maps.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+#| fig-align: "center"
+#| fig-cap: "Map of the City of Toronto Showing its 25 Wards"
+knitr::include_graphics("../other/images/toronto_wards.png")
+```
 
@@
 # Read in the cleaned demographic data
-demographic_data <- read_csv("../data/analysis_data/clean_ward_demo_data.csv")
+demographic_data <- read_csv("../data/analysis_data/clean_ward_data.csv")
 
@@
   theme(
-    plot.title = element_text(size = 16, hjust = 0.5),
+    plot.title = element_text(size = 14, hjust = 0.5),
     axis.text.x = element_text(angle = 45, hjust = 1, size = 10),

### 2024-09-27 paper/paper.qmd
updated paper.pdf with final changes
- updated position on figures
- small grammatical changes

@@
 ```
+
+\newpage
 
@@
 
--   Per Capita Funding: Total capital funding allocated to a ward divided by its population.
+-   `Per Capita Funding`: Total capital funding allocated to a ward divided by its population.
 
--   Median Household Income: The median income of households within a ward.
+-   `Median Household Income`: The median income of households within a ward.
 
--   Low-Income Prevalence: The percentage of households classified as low-income in a ward.
+-   `Low-Income Prevalence`: The percentage of households classified as low-income in a ward.
 
--   Educational Attainment: The percentage of residents with a bachelor's degree or higher.
+-   `Educational Attainment`: The percentage of residents with a bachelor's degree or higher.
 
--   Visible Minority Population: The percentage of residents who identify as visible minorities.
+-   `Visible Minority Population`: The percentage of residents who identify as visible minorities.
 
@@
 This is because, the 2022-2031 budget period aligns more closely with the 2021 Ward Profiles, ensuring that the socioeconomic variables and funding allocations are analyzed within a similar timeframe.
+
+\newpage
 
@@
 
-\newpage
-
 ```{r Figure 1, echo=FALSE, message=FALSE, warning=FALSE}
@@
 #| fig-cap: Bar graph showing Per Capita Capital Funding by Ward
+#| fig-pos: H
 
@@
   theme(
-    plot.title = element_text(size = 14, hjust = 0.5),
-    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
-    axis.title = element_text(size = 14),
-    axis.text.y = element_text(size = 10)
+    plot.title = element_text(size = 12, hjust = 0.5),
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
+    axis.title = element_text(size = 10),
+    axis.text.y = element_text(size = 8)
   ) +
@@
 Higher income wards will be categorized as one with a median household income above \$84,000.
+
 A positive relationship was observed between per capita funding and median household income as wards above the citywide average \$84,000 generally received more funding per capita.
@@
 However, among the bottom five wards, four have median incomes below \$84,000.
-
-More specifically, Ward 2, with a high median income of \$100,000, received the second-highest per capita funding at \$4,166.43.
+Ward 2, with a high median income of \$100,000, received the second-highest per capita funding at \$4,166.43.
 In contrast, Ward 22, which has a lower median income of \$77,000, received the second-lowest per capita funding at just \$192.44.
@@
 #| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Median Household Income
+#| fig-pos: H
 
@@
 #| fig-cap: Per capita funding and low income prevalence for top and bottom 5 wards excluding ward 13 
+#| fig-pos: H
 
@@
 
-We will also classify wards where at least 33% of residents hold a bachelor's degree or higher as having high educational attainment.
+We will classify wards where at least 33% of residents hold a bachelor's degree or higher, as having high educational attainment.
 There is a clear positive relationship between educational attainment and per capita funding as wards with higher education levels tend to receive more funding.
@@
 #| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Education Levels
+#| fig-pos: H
 
@@
 #| fig-cap: Scatter plot illustrating the relationship between Per Capita Funding and Minority Population
+#| fig-pos: H
 
